\section{Problem Definition}
\label{sec:problem}

\subsection{The Communication Gap}

Mermaid~\cite{mermaid2023} has become ubiquitous. With 88{,}000+ GitHub stars, native rendering on GitHub, GitLab, Notion, and Obsidian, and 22 diagram types covering flowcharts to sequence diagrams to Gantt charts, it is the de facto standard for diagrams-as-code. Developers, architects, and technical writers reach for Mermaid daily.

Yet Mermaid's output is \emph{ugly}. Functional, yes---but never presentation-grade. The diagrams look like documentation artifacts: fixed styles, poor typography, no coherent design system, no theming beyond color overrides. Meanwhile, AI agents generating diagrams have no structured, deterministic target---they must emit raw text DSL and hope for the best.

The gap is now two-dimensional:

\begin{description}
  \item[Aesthetics gap] Mermaid is ubiquitous but visually mediocre. Teams paste Mermaid output into READMEs but would never put it in a board deck, a technical blog, or a client deliverable.
  
  \item[Agent gap] AI agents can generate structured plans, architectures, and specifications---but have no widely-adopted deterministic compiler target that accepts structured IR and guarantees byte-identical, beautiful output.
\end{description}

This compiler closes both gaps simultaneously: a \textbf{full Mermaid superset} that parses every existing Mermaid file, produces dramatically better-looking output, and provides a structured IR path for agents.

\subsection{Why Mermaid Is Necessary But Insufficient}

\subsubsection{Mermaid's Strengths (Why We Build On It)}

Mermaid~\cite{mermaid2023} earned its dominance through genuine strengths:

\begin{itemize}
  \item \textbf{Zero-install rendering} on GitHub, GitLab, Notion, Obsidian, and hundreds of platforms
  \item \textbf{Simple, memorable syntax} that developers learn in minutes
  \item \textbf{Broad coverage} --- 22 diagram types spanning flowcharts, UML, charts, and more
  \item \textbf{Git-friendly plain text} that diffs cleanly and lives alongside code
  \item \textbf{Ecosystem momentum} --- AI tools already generate Mermaid syntax
\end{itemize}

These strengths make Mermaid the right \emph{starting point}. We do not compete with Mermaid's ecosystem; we \textbf{subsume} it.

\subsubsection{Mermaid's Limitations (Why We Exist)}

\begin{itemize}
  \item \textbf{Ugly, fixed output.} No design system. Poor typography. Diagrams look like 2015 documentation tools.
  \item \textbf{No real theming.} Color overrides only; no typographic, geometric, or spatial control.
  \item \textbf{No structured IR.} Agents must generate text DSL; no JSON/YAML path for constrained generation.
  \item \textbf{No composition.} Cannot combine diagrams into multi-panel posters or referenced layouts.
  \item \textbf{Non-deterministic edge cases.} Layout varies across versions and renderers.
  \item \textbf{No animation.} Static-only output; no declarative motion.
  \item \textbf{Limited UML.} Sequence and class exist but look dated; no modern software-diagram aesthetic.
\end{itemize}

\subsubsection{The Agent Blindspot}

Modern AI agents (LLMs with tool use, planning agents, copilots) can generate structured
specifications---architectures, state machines, data models, workflows. But their options
for visual output are:

\begin{itemize}
  \item Generate raw Mermaid text (fragile; no constrained generation; ugly output)
  \item Generate SVG directly (unreliable; no semantics; no theming)
  \item Call a design tool API (proprietary; non-deterministic; heavyweight)
\end{itemize}

What agents need: a \textbf{structured IR} with a published JSON Schema, grammar-constrained
generation support, deterministic rendering, and beautiful output. That is the structured IR
path of this compiler (Part~\ref{part:frontend}).

\subsection{Who This Is For}

This compiler serves two complementary audiences through two input paths:

\subsubsection{Human Authors (Mermaid-Superset DSL)}

\begin{description}
  \item[Software engineers and architects] --- Use Mermaid daily for flowcharts, sequence diagrams, and architecture docs. Want better-looking output without changing workflow.
  
  \item[Technical writers and DevRel] --- Need diagrams for blogs, documentation, and presentations that look professional, not like auto-generated documentation artifacts.
  
  \item[Product managers and consultants] --- Need roadmaps and architecture visuals suitable for client deliverables and executive communication.
\end{description}

\subsubsection{AI Agents (Structured IR)}

AI agents require:
\begin{itemize}
  \item A well-defined intermediate representation (IR) they can generate via constrained decoding
  \item Deterministic rendering --- the same IR always produces the same visual
  \item Theming separation --- agents produce structure, not style
  \item Schema validation --- agents can verify their output conforms before rendering
  \item Embeddability --- agents can invoke rendering as a tool (MCP server)
\end{itemize}

\subsection{What Success Looks Like}

\subsubsection{Scenario: Drop-in Mermaid Upgrade}

A developer has 50 Mermaid diagrams in their documentation repository. They switch the renderer to this compiler. Every diagram parses correctly and produces output that is visually superior---better typography, more coherent layout, proper spacing---with zero file changes.

\subsubsection{Scenario: Agent-Generated Architecture Diagram}

An AI coding agent analyzes a codebase and generates a C4 component diagram as structured YAML. The compiler renders it to a presentation-grade SVG with proper stereotypes, boundaries, and relationship labels---suitable for a design review or architecture decision record.

\subsubsection{Scenario: Multi-Diagram Poster}

A technical writer composes a ``system overview'' poster: a flowchart, sequence diagram, and deployment tree in a 2$\times$2 grid with a title bar. A single \texttt{poster} file produces a cohesive multi-panel visualization.

\subsubsection{Success Criteria}

The compiler succeeds when:

\begin{enumerate}
  \item Every valid Mermaid file parses and renders correctly (Mermaid-complete)
  \item Output is demonstrably more beautiful than Mermaid's for every diagram type
  \item AI agents can generate valid structured IR via constrained decoding
  \item The same IR, rendered with the same theme, produces byte-identical output across runs
  \item Users need zero migration effort---existing Mermaid files just work
  \item The UML/software diagram line rivals or exceeds PlantUML's coverage with modern aesthetics
\end{enumerate}

\subsection{What This Is Not}

To prevent scope creep and misunderstanding:

\begin{itemize}
  \item \textbf{Not a Mermaid fork} --- we do not modify Mermaid's codebase; we are an independent compiler that accepts Mermaid syntax as input
  \item \textbf{Not a project management tool} --- does not track work, assign resources, or compute schedules
  \item \textbf{Not a design tool} --- does not provide a canvas, drag-and-drop, or WYSIWYG editing
  \item \textbf{Not interactive} --- output is static (SVG, PNG, PDF); interactivity is out of scope
  \item \textbf{Not a charting library} --- though we render charts, we are not competing with D3/Plotly for general data visualization
\end{itemize}

This is a \emph{diagram compiler}: text or structured input in, beautiful deterministic visual output out. The value is in the combination of Mermaid compatibility + superior aesthetics + determinism + agent IR + composition.
