\section{LaTeX Integration}
\label{sec:latex-integration}

Triton diagrams can be embedded in \LaTeX{} documents as \emph{vector} PDF
figures. The headline authoring mode is an \textbf{inline \texttt{triton}
environment}: you write the diagram \emph{source} directly in the \texttt{.tex}
--- exactly as you would a \texttt{tikzpicture} --- and it renders to a vector PDF
at compile time. The integration lives in its own satellite package,
\texttt{latex/} (\texttt{@triton/latex}), and is purely additive: it changes no
core behaviour and adds \textbf{zero} dependencies to the \texttt{triton} package.

\subsection{The format gap}

\LaTeX{}'s \texttt{\textbackslash includegraphics} ingests PDF, PNG, or JPG ---
never SVG --- on every engine (pdf\LaTeX{}, Xe\LaTeX{}, Lua\LaTeX{}). Triton emits
SVG (\S\ref{sec:scene}: \texttt{renderSync} $\rightarrow$ \texttt{renderSVG}). The
integration therefore converts each diagram to a vector PDF and includes
\emph{that}. It does so in one of two ways: \emph{inline} at compile time (the
default, via shell-escape), or \emph{precompiled} ahead of time and committed ---
the fallback for Overleaf, where no Node or Triton toolchain is available at
compile time.

\subsection{Inline authoring: the \texttt{triton} environment}

The \texttt{triton} environment lets you author a diagram where it appears,
with no separate render step and no figure files to manage:

\begin{lstlisting}[language=triton]
\usepackage{triton}
...
\begin{triton}
flowchart LR
  A[Start] --> B{Choice}
  B -->|yes| C[Ship it]
  B -->|no|  D[Fix it]
\end{triton}
\end{lstlisting}

When \LaTeX{} reaches \texttt{\textbackslash begin\{triton\}}, the package (a) captures the
body \emph{verbatim} via \texttt{fancyvrb}'s \texttt{VerbatimOut} to a temp file;
(b) content-hashes that source with \texttt{pdftexcmds} (\texttt{\textbackslash pdf@filemdfivesum});
(c) shells out (\texttt{\textbackslash write18}) to the \texttt{triton-latex} CLI to render a
vector PDF into a cache directory --- but \emph{only} when that hash is not
already cached, so unchanged diagrams are never re-rendered; and (d)
\texttt{\textbackslash includegraphics} the result. Because rendering is keyed on the source
hash, recompiles are cheap and deterministic.

This requires \texttt{-{}-shell-escape} (see Caveats). For Overleaf and other
locked-down compiles, the precompiled \texttt{\textbackslash triton\{name\}} command (below) is
the fallback: it includes a committed PDF and needs no toolchain at compile time.
The two share the name \texttt{triton} and dispatch on context
(\texttt{\textbackslash @currenvir}): inside \texttt{\textbackslash begin\{triton\}\ldots\textbackslash end\{triton\}} it
authors inline; used as a bare \texttt{\textbackslash triton[opts]\{name\}} it includes a
precompiled PDF.

\subsection{SVG \texorpdfstring{$\rightarrow$}{->} vector PDF, pure-JS}

The converter is pure JavaScript with \textbf{no system binaries} (no Inkscape, no
\texttt{rsvg-convert}, no headless browser). It feeds Triton's SVG through
\texttt{pdfkit} + \texttt{svg-to-pdfkit}, producing a single-page PDF whose page
box equals the diagram's \texttt{viewBox}. The output is genuinely vector:

\begin{itemize}
  \item shapes become PDF path operators (\texttt{re}, \texttt{l}, \texttt{c});
  \item the \texttt{Scene}'s five element types and the arrowhead
        \texttt{<marker>} definitions all replay faithfully;
  \item text becomes real vector glyphs in the embedded base-14 fonts
        (Helvetica / Times / Courier) --- so there is \emph{no font drift}
        between author, CI, and Overleaf, and no external font file to ship.
\end{itemize}

The SMIL animation overlays (\texttt{march}, \texttt{particle}) are dropped: a PDF
is static, and the first frame is the correct print representation.

\subsection{The CLI}

The package ships a \texttt{triton-latex} command (esbuild-bundled from
\texttt{latex/src/cli.ts}, which imports the compiler by relative path from
\texttt{../src} --- the same satellite pattern as the VS Code extension):

\begin{lstlisting}[language=ts, basicstyle=\small\ttfamily]
triton-latex render <input.triton|.mmd> -o <out.pdf|.svg> [--theme N] [--scale S]
triton-latex render-dir <srcDir>        -o <outDir>        [--theme N] [--scale S]
\end{lstlisting}

\texttt{render} converts one source to vector PDF (or passes SVG through);
\texttt{render-dir} batch-renders every \texttt{*.triton}/\texttt{*.mmd} in a
directory to \texttt{<name>.pdf}. Both reuse the core \texttt{renderSync()} path
and surface its \texttt{Result} as a clean exit code.

\subsection{The \texttt{triton.sty} package}

The \LaTeX{} package provides both the inline \texttt{triton} environment and the
precompiled-include macros. It depends only on \texttt{graphicx},
\texttt{fancyvrb}, and \texttt{pdftexcmds}, so it is engine-agnostic
(pdf\LaTeX{} / Xe\LaTeX{} / Lua\LaTeX{} / tectonic). The precompiled path resolves a
diagram by name against a configurable directory:

\begin{lstlisting}[language=ts]
\usepackage{triton}
\tritondir{figures}                 % where the rendered <name>.pdf live
\triton{flowchart}                  % includegraphics figures/flowchart.pdf
\tritonfig[width=0.6\linewidth]{avl}{An AVL tree rendered by Triton.}
\end{lstlisting}

\subsection{Caveats and limitations}

Two limitations are worth stating plainly so the inline mode is used correctly.

\paragraph{No optional \texttt{[width=\ldots]} on \texttt{\textbackslash begin\{triton\}}.}
A verbatim environment cannot carry an inline optional argument: peeking for a
\texttt{[} on the \texttt{\textbackslash begin\{triton\}} line tokenises the line break that
\texttt{fancyvrb} needs to find the end of that line, which swallows the diagram's
\emph{first} source line. (\texttt{minted} avoids this only because of its
mandatory \texttt{\{language\}} argument, which absorbs the lookahead.) So
\texttt{\textbackslash begin\{triton\}[width=\ldots]} is \textbf{not} supported. Set the
\texttt{\textbackslash includegraphics} options just \emph{before} the environment with
\texttt{\textbackslash tritonnext\{width=\ldots\}} (applies to the next diagram only), or change
the default globally with \texttt{\textbackslash tritonsetup\{width=\ldots\}}.

\paragraph{\texttt{-{}-shell-escape} is required for inline rendering.}
The inline environment shells out, so it needs shell-escape enabled and both
Node and the \texttt{triton-latex} CLI available at compile time
(\texttt{pdflatex -shell-escape}, or \texttt{tectonic -Z shell-escape}). If
shell-escape is off or the CLI is missing, the environment raises a clear
\texttt{\textbackslash PackageError} stating exactly what to fix --- it never silently emits a
blank. Overleaf disallows arbitrary shell-escape, so there you must use the
precompiled \texttt{\textbackslash triton\{name\}} path instead.

\subsection{Workflow and portability}

In the inline workflow there is no separate render step: author the source in the
\texttt{triton} environment, compile with \texttt{-{}-shell-escape}, and the
cached vector PDF is regenerated only when the source changes. This is how this
very document is built --- its figures are authored inline and rendered by Triton
at compile time (\texttt{tectonic -Z shell-escape}). For locked-down targets the
precompile loop is the fallback: render sources to PDF, commit them,
include by name with \texttt{\textbackslash triton\{name\}}, and regenerate when a diagram
changes. Because those assets are committed vector PDFs, the project compiles
unchanged on macOS, Linux, Windows, and Overleaf --- the engine axis is
irrelevant, since the asset is already a PDF by the time
\texttt{\textbackslash includegraphics} runs.

\begin{center}
\begin{tabular}{ll}
\toprule
Path & Contents \\
\midrule
\texttt{latex/src/cli.ts}   & the \texttt{triton-latex} CLI (render / render-dir) \\
\texttt{latex/src/pdf.ts}   & SVG $\rightarrow$ vector PDF (pdfkit + svg-to-pdfkit) \\
\texttt{latex/triton.sty}   & the \texttt{triton} environment + \texttt{\textbackslash triton} / \texttt{\textbackslash tritonfig} macros \\
\texttt{latex/esbuild.mjs}  & bundles the CLI + compiler graph from \texttt{src/} \\
\texttt{latex/examples/}    & \texttt{inline-demo.tex} (inline env), plus precompile sources / \texttt{figures/} PDFs \\
\bottomrule
\end{tabular}
\end{center}
