\section{Composition: Multi-Panel Posters}
\label{sec:composition}

The Composition layer arranges multiple grammar outputs into a single poster or infographic. This is how the multi-section corpus images (10x-coding-playbook, archon-harness-builder, rag-vectorless-explainer) are assembled.

\paragraph{Implementation Status.}
Fully implemented in \texttt{packages/core/src/composition/} with grid-based layout and \texttt{scene-transform.ts} kernel helper (\texttt{translateAndScale}, \texttt{embedSceneInRect}). Supports content-driven row sizing, multi-cell posters, and inline grammar IR embedding. CompositionTheme with light and dark-poster variants. \textbf{Deferred:} \texttt{ir\_file} external URI references (pkg:, file:, http:); currently inline embedding only.

% =========================================================================
\subsection{Design Philosophy}
% =========================================================================

\paragraph{Composition is a thin layer.}
The Composition layer does not define new visual primitives. It arranges existing grammar outputs (each a Scene IR) into a larger canvas.

\paragraph{Each panel is an independent grammar.}
A panel might be a timeline, a flow diagram, a sequence diagram, a tree, or a simple content element (stat callout, text block). Each panel has its own Domain IR and is compiled independently.

\paragraph{The composition specifies arrangement, not content.}
Content is defined in each panel's Domain IR. The Composition IR specifies panel positions, sizes, and inter-panel elements (dividers, titles).

\paragraph{Grammar\,=\,semantics; theme\,=\,style applies to composition too.}
The Composition IR describes WHAT is arranged WHERE (structure). HOW it looks (borders, backgrounds, title fonts, gap fills) is a \texttt{CompositionTheme} --- never embedded in the IR.

% =========================================================================
\subsection{Definitions}
\label{sec:composition-defs}
% =========================================================================

\begin{description}
  \item[Composition] A document that holds one or more \textbf{cells} arranged in a deterministic grid layout.  A Composition compiles to a single merged Scene IR.
  \item[Cell] A positioned slot in the grid.  A cell either (a) references/embeds a sub-document of any grammar (timeline, sequence, tree, flow), or (b) holds a simple content element (title block, stat callout, text block, image).
  \item[Grid] The canonical layout model: rows $\times$ columns.  A grid of $R$ rows and $C$ columns defines $R \times C$ unit slots.  Cells may span multiple rows or columns.  A \emph{stack} is the special case $C=1$ (vertical) or $R=1$ (horizontal).
  \item[Panel Chrome] Visual decoration around/within a cell (border, background, title bar, caption) rendered by the composition engine from theme tokens.
  \item[Sub-Scene] The Scene IR produced by compiling a cell's grammar content through that grammar's layout engine.
\end{description}

% =========================================================================
\subsection{Composition IR --- Formal Shape}
\label{sec:composition-ir}
% =========================================================================

The Composition IR is a small, total, self-contained document.  Every field has a well-defined type; optional fields state their defaults.

\begin{lstlisting}[language=yaml,caption={Composition IR --- complete type definition (pseudo-schema)}]
CompositionDocument:
  version: "1.0"                      # REQUIRED; semver-major
  metadata:                           # REQUIRED
    title: string                     # REQUIRED; poster title
    subtitle?: string                 # default: absent
    theme?: string                    # CompositionTheme name; default: "default"
    canvas:
      width: number                   # REQUIRED; logical px
      height?: number | "auto"        # default: "auto" (computed)
  grid:                               # REQUIRED; canonical layout
    columns: integer                  # >= 1; REQUIRED
    rows?: integer | "auto"           # default: "auto" (ceil(cellCount / columns))
    gap?: number                      # inter-cell gap in px; default: 0
    padding?: number                  # outer canvas padding; default: 0
  cells: Cell[]                       # REQUIRED; >= 1 cell; row-major order

Cell:
  id: string                          # REQUIRED; unique within composition
  row?: integer                       # 0-indexed; default: assigned row-major
  col?: integer                       # 0-indexed; default: assigned row-major
  rowSpan?: integer                   # default: 1
  colSpan?: integer                   # default: 1
  title?: string                      # cell title; rendered as panel chrome
  caption?: string                    # below-content caption
  content: CellContent                # REQUIRED; what this cell holds

CellContent:                          # discriminated union on "kind"
  | GrammarContent                    # sub-diagram (any grammar)
  | StatContent                       # stat callout (number + caption)
  | TextContent                       # rich text block
  | TitleContent                      # section title / heading
  | ImageContent                      # static image embed

GrammarContent:
  kind: "grammar"
  grammar: "timeline" | "sequence" | "tree" | "flow"   # grammar name
  ir?: object                         # inline Domain IR (grammar-specific)
  ir_file?: string                    # OR external file reference
  # Exactly one of ir / ir_file REQUIRED.

StatContent:
  kind: "stat"
  value: string                       # e.g. "10x", "99.9%"
  label: string                       # description beneath stat
  trend?: "up" | "down" | "neutral"   # optional trend indicator

TextContent:
  kind: "text"
  body: string                        # markdown-subset (bold, italic, bullet)
  align?: "left" | "center" | "right" # default: "left"

TitleContent:
  kind: "title"
  text: string
  level?: 1 | 2 | 3                  # heading level; default: 1

ImageContent:
  kind: "image"
  src: string                         # file path or data URI
  alt?: string                        # accessibility text
\end{lstlisting}

\paragraph{Identity.}
Each \texttt{Cell.id} is unique within a composition.  \texttt{CompositionDocument.metadata.title} serves as the document identity for hashing and golden tests.

\paragraph{Row-major default ordering.}
If \texttt{row}/\texttt{col} are omitted from cells, they are assigned left-to-right, top-to-bottom in declaration order.  Explicit placement overrides the default.

\paragraph{Invariants.}
\begin{itemize}
  \item All \texttt{Cell.id} values are unique.
  \item Cell placements must not overlap (no two cells occupy the same grid slot).
  \item Spans must not exceed grid dimensions: \texttt{col + colSpan <= columns}, \texttt{row + rowSpan <= rows}.
  \item Exactly one of \texttt{ir} or \texttt{ir\_file} must be present in \texttt{GrammarContent}.
\end{itemize}

% =========================================================================
\subsection{The Sub-Scene Embed Mechanism}
\label{sec:composition-embed}
% =========================================================================

This section specifies \textbf{precisely} how independently-compiled sub-scenes are embedded into one merged Scene.  This is the core composition mechanism.

\subsubsection{Step 1: Compile Each Cell's Content to a Sub-Scene}

For each cell whose content is \texttt{kind: "grammar"}:
\begin{enumerate}
  \item Resolve the Domain IR (inline \texttt{ir} or load \texttt{ir\_file}).
  \item Validate via that grammar's schema.
  \item Invoke the grammar's layout engine: e.g., \texttt{buildSequenceScene(ir, theme)} $\to$ \texttt{Scene}.
  \item The resulting Scene has deterministic \texttt{width}, \texttt{height}, and \texttt{primitives[]}.
\end{enumerate}

For content elements (\texttt{stat}, \texttt{text}, \texttt{title}, \texttt{image}): a small content-layout function produces a Scene of known dimensions (text-measured).

Each cell now owns a \textbf{sub-scene} $S_i$ with dimensions $(w_i, h_i)$.

\subsubsection{Step 2: Compute Grid Layout (Cell Rectangles)}

The composition engine computes a deterministic rectangle $(x, y, W, H)$ for each cell based on the grid, sub-scene sizes, gaps, and padding.  The algorithm is specified in Section~\ref{sec:composition-layout-contract}.

\subsubsection{Step 3: Translate and Scale Sub-Scenes into Cell Rectangles}

For each cell $i$ with sub-scene $S_i$ (dimensions $w_i \times h_i$) and allocated rectangle $(x_i, y_i, W_i, H_i)$:

\begin{enumerate}
  \item \textbf{Compute scale factor} (uniform, aspect-preserving):
    \[
      s_i = \min\!\left(\frac{W_i}{w_i},\; \frac{H_i}{h_i}\right)
    \]
    If $s_i \geq 1.0$ (sub-scene fits without scaling), set $s_i = 1.0$ (no upscale).

  \item \textbf{Compute centering offset} within the cell rectangle:
    \[
      \Delta x_i = x_i + \frac{W_i - s_i \cdot w_i}{2}, \qquad
      \Delta y_i = y_i + \frac{H_i - s_i \cdot h_i}{2}
    \]

  \item \textbf{Apply transform to every primitive} in $S_i.\text{primitives}$:
    \[
      \texttt{translateAndScale}(p,\; \Delta x_i,\; \Delta y_i,\; s_i) \to p'
    \]
    This is a pure function over Scene primitives (definition below).

  \item \textbf{Collect} all transformed primitives $p'$ into the output list.
\end{enumerate}

\paragraph{The \texttt{translateAndScale} kernel helper.}
\label{para:translate-scale-helper}

\textbf{FLAG FOR BARBARA/MARK:} The composition layer requires a pure utility function on Scene primitives:

\begin{lstlisting}[language=javascript,caption={translateAndScale --- kernel helper (specification)}]
/**
 * Pure function: offsets and uniformly scales all coordinate fields of a
 * ScenePrimitive (and recursively, GroupPrimitive children).
 *
 * Coordinates: (x,y) -> (x*s + dx, y*s + dy)
 * Dimensions:  (w,h) -> (w*s, h*s)
 * Font sizes:  fontSize -> fontSize * s
 * Stroke widths: strokeWidth -> strokeWidth * s
 * Radii: r, rx -> r*s, rx*s
 *
 * Path `d` strings: all numeric coordinates in the path are scaled/translated.
 * StrokeGradient coordinates: x1,y1,x2,y2 transformed identically.
 * GroupPrimitive: recurse into children.
 *
 * Determinism: output rounded via rhu(2dp) per the determinism contract.
 */
function translateAndScale(
  primitive: ScenePrimitive,
  dx: number, dy: number,
  scale: number
): ScenePrimitive;

/**
 * Convenience: transform an entire Scene into a target rectangle.
 * Returns a new primitives[] array (the Scene itself is not mutated).
 */
function embedSceneInRect(
  scene: Scene,
  targetRect: { x: number; y: number; width: number; height: number }
): ScenePrimitive[];
\end{lstlisting}

This function lives in the kernel (e.g., \texttt{packages/core/src/scene-transform.ts}) because it operates on Scene IR primitives directly.  It must handle \emph{every} primitive kind defined in \texttt{scene.ts}: Line, Rect, Circle, Text, MultiText, Path, Group, Image.

\subsubsection{Step 4: Render Panel Chrome}

After sub-scenes are embedded, the composition engine emits additional Scene primitives for chrome:
\begin{itemize}
  \item Cell borders (Rect primitives with stroke, no fill)
  \item Cell titles (Text primitives positioned above/inside cell rect)
  \item Cell captions (Text primitives below cell rect)
  \item Cell backgrounds (Rect primitives behind sub-scene content)
  \item Composition header/footer (Text, Image primitives at canvas top/bottom)
  \item Dividers between rows/cells (Line primitives)
\end{itemize}

All chrome styling (colors, fonts, border radii, etc.) comes from \texttt{CompositionTheme} tokens --- never from the IR.

\subsubsection{Step 5: Merge into Final Scene}

\begin{lstlisting}[language=javascript,caption={Final merge (pseudo-code)}]
function buildCompositionScene(doc: CompositionDocument): Scene {
  const grid = computeGridLayout(doc);     // Step 2
  const allPrimitives: ScenePrimitive[] = [];

  // Background (from theme)
  allPrimitives.push(backgroundRect(grid.canvasWidth, grid.canvasHeight));

  // Header chrome
  allPrimitives.push(...renderHeader(doc.metadata, theme));

  // Cells: chrome-behind, then sub-scene, then chrome-above
  for (const cell of grid.cells) {  // row-major order
    allPrimitives.push(...renderCellBackground(cell, theme));
    allPrimitives.push(...embedSceneInRect(cell.subScene, cell.rect));
    allPrimitives.push(...renderCellChrome(cell, theme));  // borders, title
  }

  // Footer chrome
  allPrimitives.push(...renderFooter(doc.metadata, theme));

  return {
    width: grid.canvasWidth,
    height: grid.canvasHeight,
    background: theme.canvasBackground,
    primitives: allPrimitives,
  };
}
\end{lstlisting}

\paragraph{Determinism.}
The merged Scene is deterministic because: (a) each sub-scene is deterministic (grammar contract), (b) cell iteration is row-major (fixed order), (c) \texttt{translateAndScale} is a pure function with rhu-rounded output, (d) chrome rendering is theme-deterministic.  The resulting Scene can be hashed via \texttt{sceneHash()} for golden testing.

% =========================================================================
\subsection{Layout Contract --- Deterministic Grid Sizing}
\label{sec:composition-layout-contract}
% =========================================================================

The grid sizing algorithm computes exact cell rectangles from sub-scene dimensions, spans, gaps, and padding.  It is fully deterministic (no iteration, no heuristics).

\subsubsection{Inputs}

\begin{itemize}
  \item $C$ = number of columns (from \texttt{grid.columns})
  \item $R$ = number of rows (explicit or $\lceil |\text{cells}| / C \rceil$)
  \item $G$ = gap (from \texttt{grid.gap}, default 0)
  \item $P$ = padding (from \texttt{grid.padding}, default 0)
  \item $W_{\text{canvas}}$ = canvas width (from \texttt{metadata.canvas.width})
  \item For each cell $i$: sub-scene dimensions $(w_i, h_i)$, placement $(r_i, c_i)$, spans $(\text{rs}_i, \text{cs}_i)$
\end{itemize}

\subsubsection{Column Width Computation}

Available width after padding and gaps:
\[
  W_{\text{avail}} = W_{\text{canvas}} - 2P - (C-1) \cdot G
\]

Column widths are computed as \textbf{content-driven maximum}:
\[
  \text{colWidth}[j] = \max_{\substack{i \,:\, c_i = j,\\ \text{cs}_i = 1}} w_i
  \qquad \text{(max natural width of single-span cells in column } j\text{)}
\]

If the sum of column widths exceeds $W_{\text{avail}}$, columns are \textbf{proportionally scaled} to fit:
\[
  \text{colWidth}[j] \leftarrow \text{colWidth}[j] \cdot \frac{W_{\text{avail}}}{\sum_k \text{colWidth}[k]}
\]

Multi-column-span cells: their natural width is distributed across spanned columns only if needed (max propagation).

\subsubsection{Row Height Computation}

Row heights are computed analogously:
\[
  \text{rowHeight}[r] = \max_{\substack{i \,:\, r_i = r,\\ \text{rs}_i = 1}} h_i
  \qquad \text{(max natural height of single-span cells in row } r\text{)}
\]

Row heights are \emph{not} constrained by canvas height (canvas height is derived from rows when \texttt{height: "auto"}).

\subsubsection{Cell Rectangle Computation}

For cell $i$ at grid position $(r_i, c_i)$ with spans $(\text{rs}_i, \text{cs}_i)$:

\begin{align}
  x_i &= P + \sum_{j=0}^{c_i - 1} (\text{colWidth}[j] + G) \\
  y_i &= P + H_{\text{header}} + \sum_{k=0}^{r_i - 1} (\text{rowHeight}[k] + G) \\
  W_i &= \sum_{j=c_i}^{c_i + \text{cs}_i - 1} \text{colWidth}[j] + (\text{cs}_i - 1) \cdot G \\
  H_i &= \sum_{k=r_i}^{r_i + \text{rs}_i - 1} \text{rowHeight}[k] + (\text{rs}_i - 1) \cdot G
\end{align}

Where $H_{\text{header}}$ accounts for the composition header chrome height (if present).

\subsubsection{Canvas Height (auto)}

\[
  H_{\text{canvas}} = 2P + H_{\text{header}} + H_{\text{footer}} + \sum_{k=0}^{R-1} \text{rowHeight}[k] + (R-1) \cdot G
\]

\paragraph{Determinism guarantee.}
All operations are max, sum, and proportional scaling --- no iteration, no convergence, no randomness.  The same input always produces the same rectangles.

% =========================================================================
\subsection{Determinism and Theme/Semantics Split}
\label{sec:composition-theme}
% =========================================================================

\subsubsection{What the IR Carries (Semantics)}

\begin{itemize}
  \item Grid structure: columns, rows, cell placement, spans
  \item Cell content: grammar type + domain IR (what to draw)
  \item Cell identity: id, title, caption (semantic labels)
  \item Composition metadata: title, subtitle
\end{itemize}

\subsubsection{What the Theme Carries (Style)}

\begin{lstlisting}[language=javascript,caption={CompositionTheme type (specification)}]
interface CompositionTheme {
  // Canvas
  canvasBackground: string;       // CSS color

  // Grid spacing (override IR defaults when theme wants control)
  gap?: number;                   // overrides grid.gap if set
  padding?: number;               // overrides grid.padding if set

  // Cell chrome
  cellBorder: { color: string; width: number; radius: number };
  cellBackground: string;         // per-cell fill (or "transparent")
  cellTitleFont: { family: string; size: number; weight: number; color: string };
  cellCaptionFont: { family: string; size: number; weight: number; color: string };
  cellTitlePosition: "above" | "inside-top";

  // Header/Footer
  headerFont: { family: string; size: number; weight: number; color: string };
  subtitleFont: { family: string; size: number; weight: number; color: string };
  footerBadge: { fill: string; textColor: string; radius: number };

  // Dividers
  dividerColor: string;
  dividerThickness: number;

  // Scale policy
  scalePolicy: "fit" | "clip" | "overflow";  // how oversized sub-scenes handled
}
\end{lstlisting}

\paragraph{The boundary.}
The IR says ``cell A1 holds a flow diagram titled `Pipeline'.''  The theme says ``render that title in Inter 18px bold white, with a 1px \#333 border and 12px corner radius.''  This separation ensures the same composition IR renders differently under different themes (dark-poster, light-dashboard, print) without IR modification.

% =========================================================================
\subsection{Edge Cases}
\label{sec:composition-edge-cases}
% =========================================================================

\subsubsection{Sub-Scene Larger Than Cell (Scale-to-Fit)}

When a sub-scene's natural dimensions exceed its allocated cell rectangle:

\begin{itemize}
  \item \textbf{Default policy (\texttt{scalePolicy: "fit"}):} The sub-scene is uniformly scaled \emph{down} (aspect-preserving) to fit entirely within the cell.  The scale factor $s = \min(W/w, H/h)$ from Section~\ref{sec:composition-embed} handles this automatically.  The sub-scene is centered within the remaining space.
  \item \textbf{Alternative (\texttt{scalePolicy: "clip"}):} The sub-scene is not scaled; overflow is clipped at the cell boundary (implemented via a clip-rect Group primitive).
  \item \textbf{Alternative (\texttt{scalePolicy: "overflow"}):} The sub-scene is rendered at natural size; overflow is visible (overlaps adjacent cells).  Use only for artistic effect.
\end{itemize}

\textbf{Decision:} Default is \texttt{"fit"} (scale-to-fit, never upscale).  This ensures every sub-scene is fully visible and the composition is self-contained.

\subsubsection{Empty Cell}

A cell with no content (or content that produces an empty Scene with $w=0, h=0$):
\begin{itemize}
  \item The cell rectangle is allocated normally by the grid algorithm (it may still have non-zero size due to other cells in the same row/column).
  \item No sub-scene primitives are emitted for that cell.
  \item Cell chrome (border, background) is still rendered if theme specifies it.
\end{itemize}

\subsubsection{Single-Cell Composition}

A composition with one cell is valid.  It produces the sub-scene embedded in the canvas with chrome (header, footer, border).  This is useful for wrapping a single diagram in a poster frame.

\subsubsection{Ragged Grid (Missing Cells)}

If the grid has $R \times C$ slots but fewer cells are declared:
\begin{itemize}
  \item Undeclared slots are treated as empty cells (no content, no chrome).
  \item Row heights and column widths are still computed from the cells that \emph{are} present.
  \item The grid is NOT filled with phantom cells---only declared cells exist.
\end{itemize}

\subsubsection{Nested Compositions}

A cell's \texttt{GrammarContent} MAY reference \texttt{grammar: "composition"}, embedding a nested Composition document.  This is \textbf{allowed} but bounded:
\begin{itemize}
  \item Nesting depth is limited to 3 levels (configurable; validation rejects deeper nesting).
  \item The nested composition compiles to a sub-Scene like any other grammar.
  \item No special handling is needed---the embed mechanism is grammar-agnostic.
\end{itemize}

\textbf{Rationale:} Nested compositions enable complex layouts (e.g., a 2$\times$2 sub-grid within a cell of a larger grid) without introducing a richer layout primitive.

% =========================================================================
\subsection{Worked Example: Multi-Grammar Poster}
\label{sec:composition-example}
% =========================================================================

This example reproduces the structure of a corpus poster (inspired by the RAG-vectorless-explainer, sample-images/image copy 5): a $2\times 2$ grid whose cells hold a flow pipeline, a tree hierarchy, a sequence diagram, and a stat callout.

\subsubsection{Composition IR}

\begin{lstlisting}[language=yaml,caption={Worked example: RAG Architecture Poster}]
version: "1.0"
metadata:
  title: "RAG Architecture Deep Dive"
  subtitle: "Retrieval-Augmented Generation Pipeline"
  theme: dark-poster
  canvas:
    width: 1200
    height: auto
grid:
  columns: 2
  rows: 2
  gap: 24
  padding: 32
cells:
  - id: pipeline
    row: 0
    col: 0
    title: "Ingestion Pipeline"
    content:
      kind: grammar
      grammar: flow
      ir:
        direction: left-to-right
        nodes:
          - { id: docs, label: "Documents", category: input }
          - { id: chunk, label: "Chunker", category: step }
          - { id: embed, label: "Embedder", category: step }
          - { id: store, label: "Vector DB", category: output }
        edges:
          - { from: docs, to: chunk }
          - { from: chunk, to: embed }
          - { from: embed, to: store }

  - id: taxonomy
    row: 0
    col: 1
    title: "Document Taxonomy"
    content:
      kind: grammar
      grammar: tree
      ir:
        root:
          id: corpus
          label: "Corpus"
          children:
            - id: pdf
              label: "PDFs"
              children:
                - { id: papers, label: "Papers" }
                - { id: manuals, label: "Manuals" }
            - id: web
              label: "Web Pages"
              children:
                - { id: docs, label: "Docs" }
                - { id: blogs, label: "Blogs" }

  - id: retrieval
    row: 1
    col: 0
    title: "Query Flow"
    content:
      kind: grammar
      grammar: sequence
      ir:
        participants:
          - { id: user, label: "User" }
          - { id: api, label: "API" }
          - { id: vdb, label: "VectorDB" }
          - { id: llm, label: "LLM" }
        messages:
          - { from: user, to: api, label: "query", order: 1 }
          - { from: api, to: vdb, label: "embed + search", order: 2 }
          - { from: vdb, to: api, label: "top-k chunks", order: 3 }
          - { from: api, to: llm, label: "prompt + context", order: 4 }
          - { from: llm, to: api, label: "response", order: 5 }
          - { from: api, to: user, label: "answer", order: 6 }

  - id: accuracy
    row: 1
    col: 1
    title: "Retrieval Accuracy"
    content:
      kind: stat
      value: "94.2%"
      label: "Top-5 recall on evaluation set"
      trend: up
\end{lstlisting}

\subsubsection{Compilation Walkthrough}

\begin{enumerate}
  \item \textbf{Parse \& validate:} The composition IR is validated (4 cells, 2$\times$2 grid, no overlaps, all grammar IRs valid).

  \item \textbf{Compile sub-scenes:}
  \begin{itemize}
    \item Cell \texttt{pipeline}: \texttt{buildFlowScene(ir)} $\to$ Scene ($w{=}480, h{=}120$)
    \item Cell \texttt{taxonomy}: \texttt{buildTreeScene(ir)} $\to$ Scene ($w{=}360, h{=}280$)
    \item Cell \texttt{retrieval}: \texttt{buildSequenceScene(ir)} $\to$ Scene ($w{=}520, h{=}340$)
    \item Cell \texttt{accuracy}: \texttt{buildStatScene(ir)} $\to$ Scene ($w{=}200, h{=}100$)
  \end{itemize}

  \item \textbf{Grid sizing:}
  \begin{itemize}
    \item $W_{\text{avail}} = 1200 - 2(32) - 1(24) = 1112$ px
    \item Column widths (max of single-span cells): col0 $= \max(480, 520) = 520$, col1 $= \max(360, 200) = 360$
    \item Sum $= 880 < 1112$, so no scaling needed.  Columns keep natural widths (centered in available space, or expanded proportionally --- theme decision).
    \item Row heights: row0 $= \max(120, 280) = 280$, row1 $= \max(340, 100) = 340$
  \end{itemize}

  \item \textbf{Cell rectangles} (with padding $P{=}32$, gap $G{=}24$):
  \begin{itemize}
    \item \texttt{pipeline}: $(32, 32{+}H_{\text{hdr}}, 520, 280)$
    \item \texttt{taxonomy}: $(32{+}520{+}24, 32{+}H_{\text{hdr}}, 360, 280)$
    \item \texttt{retrieval}: $(32, 32{+}H_{\text{hdr}}{+}280{+}24, 520, 340)$
    \item \texttt{accuracy}: $(576, 32{+}H_{\text{hdr}}{+}280{+}24, 360, 340)$
  \end{itemize}

  \item \textbf{Embed sub-scenes:}
  \begin{itemize}
    \item \texttt{pipeline} ($480{\times}120$ into $520{\times}280$): $s = \min(520/480, 280/120) = \min(1.08, 2.33) \to 1.0$ (no upscale).  Centered: $\Delta x = 32 + (520{-}480)/2 = 52$, $\Delta y = 32 + H_{\text{hdr}} + (280{-}120)/2$.
    \item \texttt{taxonomy} ($360{\times}280$ into $360{\times}280$): exact fit, $s=1.0$, no offset.
    \item \texttt{retrieval} ($520{\times}340$ into $520{\times}340$): exact fit.
    \item \texttt{accuracy} ($200{\times}100$ into $360{\times}340$): $s=1.0$, centered in larger rect.
  \end{itemize}

  \item \textbf{Panel chrome:} Borders, titles (``Ingestion Pipeline'', etc.), backgrounds emitted per theme.

  \item \textbf{Final Scene:} $1200 \times (2 \cdot 32 + H_{\text{hdr}} + H_{\text{ftr}} + 280 + 340 + 24)$ --- all primitives from 4 sub-scenes + chrome in a single \texttt{Scene} object.  \texttt{sceneHash()} is deterministic.
\end{enumerate}


\ourdiagram[0.85\linewidth]{rag-poster-layout}{%
  Composition structure of the RAG Architecture Deep Dive poster:
  four panels (Ingestion Pipeline, Document Taxonomy, Query Flow, Retrieval Accuracy)
  arranged in a $2{\times}2$ grid, each compiled by its grammar-specific layout engine
  and assembled into a single Unified Scene IR.}



% =========================================================================
\subsection{Panel References}
\label{sec:composition-panel-refs}
% =========================================================================

Panels can reference external IR files instead of inline definitions:

\begin{lstlisting}[language=yaml,caption={Panel with external IR reference}]
cells:
  - id: timeline-panel
    content:
      kind: grammar
      grammar: timeline
      ir_file: "./roadmap.timeline.yaml"  # external file
\end{lstlisting}

This enables:
\begin{itemize}
  \item Reusing the same diagram in multiple compositions
  \item Separate version control for each panel's IR
  \item Incremental updates (change one panel, not the whole poster)
\end{itemize}

% =========================================================================
\subsection{Compilation Pipeline Summary}
\label{sec:composition-pipeline}
% =========================================================================

\begin{enumerate}
  \item \textbf{Parse Composition IR}: Validate structure, resolve grammar references, check grid invariants.
  
  \item \textbf{Compile each cell}: For each cell, invoke the appropriate grammar's layout engine to produce a sub-Scene.
  
  \item \textbf{Compute grid layout}: Determine column widths, row heights, and cell rectangles (Section~\ref{sec:composition-layout-contract}).
  
  \item \textbf{Embed sub-scenes}: For each cell, \texttt{embedSceneInRect(subScene, cellRect)} produces translated/scaled primitives.
  
  \item \textbf{Render chrome}: Emit background, borders, titles, captions, header, footer, dividers from \texttt{CompositionTheme}.
  
  \item \textbf{Merge}: Concatenate all primitives in painter's-algorithm order (back to front) into a final Scene with computed width/height.
  
  \item \textbf{Hash}: \texttt{sceneHash(finalScene)} for golden testing.
\end{enumerate}

% =========================================================================
\subsection{Limitations}
\label{sec:composition-limitations}
% =========================================================================

\begin{description}
  \item[Cross-panel connectors (specified, not yet implemented)]
        Edges that span from a node in one panel to a node in another are
        specified as a superset-only feature in
        Section~\ref{sec:cross-diagram-links}.  The implementation requires the
        node-anchor registry prerequisite described there.  Panels remain
        visually independent until that prerequisite is shipped.

  \item[No responsive layout] Panel sizes are computed once; there is no responsive reflow for different canvas widths.
  
  \item[No interactive linking] Clicking a panel does not navigate or filter. Output is static.
  
  \item[No upscaling] Sub-scenes are never scaled up (only down or 1:1). Prevents pixelation of content designed for smaller viewports.
\end{description}

These constraints keep the Composition layer simple and deterministic.  Cross-panel
connections are addressed in Section~\ref{sec:cross-diagram-links}; responsive layout
remains out of scope.

% =========================================================================
\subsection{Open Questions}
\label{sec:composition-open-questions}
% =========================================================================

\paragraph{For Mark (Schema):}
\begin{itemize}
  \item Final JSON Schema shape for \texttt{CompositionDocument} --- should \texttt{CellContent} use a discriminated union (\texttt{kind} field) or separate top-level keys?
  \item Cell reference vs.\ inline embed: should \texttt{ir\_file} support URI schemes (e.g., \texttt{pkg:} for package-internal references)?
  \item Validation: should the composition schema embed sub-grammar schemas or validate them separately (two-pass)?
  \item Should \texttt{grid.columns}/\texttt{grid.rows} allow named tracks (like CSS Grid's named lines)?
\end{itemize}

\paragraph{For Barbara (Rendering):}
\begin{itemize}
  \item \textbf{The \texttt{translateAndScale} kernel helper} (Section~\ref{sec:composition-embed}): implement as a pure function in \texttt{packages/core/src/scene-transform.ts}.  Must handle Path \texttt{d}-string coordinate transformation and StrokeGradient coordinate transformation.
  \item Scale-to-fit policy: should the default clip oversized sub-scenes or scale them?  (Spec recommends scale-to-fit; Barbara to confirm UX.)
  \item Panel chrome rendering: confirm that cell borders/titles use existing primitives (Rect, Text) and need no new Scene IR additions.
  \item How should \texttt{scalePolicy: "clip"} be implemented?  Via a GroupPrimitive with a clip-path attribute (requires Scene IR extension) or via a post-processing step?
\end{itemize}

\paragraph{Cross-references:}
\begin{itemize}
  \item Grammar concept and two-IR model: Section~\ref{sec:grammar-concept}
  \item Scene IR primitives: Section~\ref{sec:scene-ir} and \texttt{packages/core/src/scene.ts}
  \item Flow Grammar: Section~\ref{sec:flow-grammar}
  \item Sequence Grammar: Section~\ref{sec:sequence-grammar}
  \item Tree Grammar: Section~\ref{sec:tree-grammar}
  \item Determinism contract: Section~\ref{sec:determinism}
\end{itemize}
