% =============================================================================
% §30b — Cross-Diagram Node Linking
% =============================================================================
% Specifies the mechanism for drawing edges between nodes that live in
% different cells of a poster/composition.  Superset-only.  Spec-first (2026).
%
% §30b covers two levels of cross-diagram linking:
%   • Atomic link  — a single directed or undirected overlay edge
%                    between two specific nodes in different cells.
%   • trace        — a named, ordered, optionally-typed multi-hop path
%                    (A→B→C→…) representing a logical thread woven
%                    across system layers/views.  The headline use case
%                    is layered system traceability.
% =============================================================================

\section{Cross-Diagram Node Linking}
\label{sec:cross-diagram-links}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Concept and Motivation}
% ─────────────────────────────────────────────────────────────────────────────

A poster assembled from independent grammar cells
(Section~\ref{sec:composition}) is, by default, a set of \emph{panels side by
side}: each cell compiles and renders autonomously with no structural
relationship to its neighbours.  Cross-diagram node linking removes that
constraint.  A \texttt{link} statement in the poster DSL draws a directed or
undirected edge between a node in one cell's diagram and a node in another
cell's diagram, turning the poster into a \textbf{linked multi-view system
diagram}.

This section specifies two constructs built on the same underlying machinery:

\begin{description}
  \item[Atomic \texttt{link}.] A single cross-diagram edge between a source
        node and a target node.  This is the primitive from which all
        cross-diagram overlay drawing is built.

  \item[\texttt{trace}.] A \emph{named, ordered, optionally-typed multi-hop
        path} of atomic links.  A trace represents a logical thread --- such
        as a user journey, a distributed request, or a requirements
        traceability chain --- woven across the diagram layers of a poster.
        This is the flagship use case for cross-diagram linking: representing
        system traceability across abstraction levels.  The \texttt{trace}
        construct is specified in Section~\ref{sec:cdl-trace}.
\end{description}

\begin{description}
  \item[Presentation-layer assertion.]
        A cross-diagram link does not modify either diagram's Domain IR or
        Scene IR.  It is an annotation on the composition itself, resolved
        after every cell has compiled independently.

  \item[Superset-only.]
        The \texttt{link} and \texttt{trace} statements are new top-level
        poster keywords with no Mermaid equivalent
        (Section~\ref{sec:superset-extensions}).  A standard Mermaid renderer
        will reject a poster document that contains these statements; this is
        the intended signal that the file requires our compiler.

  \item[Additive and non-fatal.]
        Unresolvable links (unknown cell, unknown node, non-linkable diagram
        type) emit a warning and are skipped.  The poster and all valid links
        still render.
\end{description}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Link Syntax}
\label{sec:cdl-syntax}
% ─────────────────────────────────────────────────────────────────────────────

A \texttt{link} statement is a poster-level declaration written alongside
\texttt{cell} definitions.  Its abstract grammar is:

\begin{lstlisting}[language=,caption={Cross-diagram link statement — abstract grammar}]
link <cellAddr>.<nodeId> <edge> <cellAddr>.<nodeId> [: "label"]

<cellAddr>  ::= "[" row "," col "]"   (* bracket form, 0-indexed *)
             |  colLetters rowNumber  (* Excel form, e.g. A1, B2, AA3 *)

<nodeId>    ::= (* the node identifier as authored in that cell's diagram source *)

<edge>      ::= "-->"                 (* directed arrow *)
             |  "---"                 (* plain undirected *)
             |  "-.-"                 (* dashed undirected *)
             |  "-.."                 (* dashed undirected, alternate *)
             |  "-.->"|"-.->"        (* dashed directed arrow *)
\end{lstlisting}

Cell addresses reuse the dual addressing scheme introduced for the
\texttt{poster} DSL and specified in Section~\ref{sec:composition-posters}:
the bracket form \texttt{[row,col]} is 0-indexed (row first); the Excel form
\texttt{A1}, \texttt{B2}, \ldots{} uses bijective base-26 column letters with
1-indexed row numbers.  Both forms may be mixed freely within a single poster.

The optional \texttt{: "label"} suffix attaches a text label rendered at
the midpoint of the overlay edge, styled by the composition theme's
\texttt{overlayLabelFont} token.

\paragraph{Worked example.}

The following poster specifies an Order Fulfilment system with four cells and
three cross-diagram links that connect nodes across cells:

\begin{lstlisting}[language=,caption={Cross-diagram links in a three-cell poster (flow, class, state)}]
---
theme: executive
layout: grid 3x1
---
poster "Order Fulfilment System"

  cell A1: flowchart LR
    recv[Receive Order] --> valid[Validate]
    valid --> pay[Payment]
    valid --> pick[Warehouse Pick]
    valid --> ship[Dispatch]

  cell B1: classDiagram
    class PaymentGateway
    class WarehouseSystem
    class OrderService
    OrderService --> PaymentGateway
    OrderService --> WarehouseSystem

  cell C1: stateDiagram-v2
    [*] --> Pending
    Pending --> Processing
    Processing --> Shipped
    Shipped --> [*]

link A1.pay --> B1.PaymentGateway : "handled by"
link A1.pick --> B1.WarehouseSystem : "fulfilled by"
link A1.ship --> C1.Shipped : "triggers"
\end{lstlisting}

\ourdiagram[\linewidth]{link-poster}{%
  Rendered poster: three cross-diagram overlay edges connect the order-flow
  cell (left) to the service-class cell (centre) and the order-state cell
  (right).  The two adjacent-cell links (\texttt{handled by},
  \texttt{fulfilled by}) route \emph{directly} through the vertical gutter
  between the flow and class cells; the non-adjacent skip-link
  (\texttt{triggers}, A1$\to$C1, which would cross the centre cell) falls back
  to the bottom routing bus.  Node positions come from the per-grammar anchor
  registries.}

\texttt{link A1.pay --> B1.PaymentGateway} reads: ``draw a directed arrow
from the node with id \texttt{pay} in cell \texttt{A1} to the class named
\texttt{PaymentGateway} in cell \texttt{B1}''.  The cell addresses use
Excel-style notation: \texttt{A1}=\texttt{[0,0]}, \texttt{B1}=\texttt{[0,1]},
\texttt{C1}=\texttt{[0,2]}.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{The Node-Anchor Registry}
\label{sec:cdl-anchor-registry}
% ─────────────────────────────────────────────────────────────────────────────

\paragraph{The core engineering prerequisite.}

Grammar layout engines today compute precise node positions internally but
\textbf{discard them}: the output of \texttt{buildFlowScene}, \texttt{buildTreeScene},
and their siblings is a bare \texttt{Scene} containing only flattened
rendering primitives.  For example, the flow layout engine builds a
\texttt{PlacedNode} record for every node (carrying \texttt{x}, \texttt{y},
\texttt{w}, \texttt{h}, \texttt{cx}, \texttt{cy}, and port coordinates) but
returns only the rendered \texttt{Scene}; there is no \texttt{nodeId
$\to$ position} registry in the output type.  \textbf{Making node positions
addressable from the composition layer is the single prerequisite that must be
resolved before any cross-diagram link can be drawn.}

\subsubsection{Candidate Implementations}

Two approaches exist for surfacing node anchors to the composition layer.

\paragraph{Candidate (a) — Populate \texttt{GroupPrimitive.id} and walk the
placed scene.}

The \texttt{GroupPrimitive} in the Scene IR carries an optional \texttt{id?:\
string} field (see \texttt{packages/core/src/scene.ts}).  Each grammar's
layout engine could wrap each node's primitives in a \texttt{GroupPrimitive}
tagged with that node's id.  At composition time the link layer would walk the
placed (already translated-and-scaled) scene, compute the axis-aligned
bounding box of each named group, and use it as the anchor.

\medskip
\textit{Pros:} No change to grammar function return types; \texttt{GroupPrimitive.id}
already exists in the IR.

\medskip
\textit{Cons:}
\begin{itemize}
  \item Recovering a bounding box requires walking the full scene tree for each
        node query (O($n$) primitives per lookup for a scene with $n$ primitives).
  \item The walk must be performed after \texttt{translateAndScale}, meaning
        coordinate reconstruction happens at the composition layer, not the
        grammar layer.
  \item No port information (which side of the node to attach an edge to) is
        recoverable from a bounding box without additional metadata.
  \item Requires every linkable grammar to wrap each node in a named
        \texttt{GroupPrimitive}, which is not currently the case and is a
        fragile convention: a grammar that emits node primitives without a
        wrapping group would silently produce empty anchors.
  \item Not reusable for hit-testing or agent node-addressing without
        additional work.
\end{itemize}

\paragraph{Candidate (b) — Sidecar \texttt{anchors} record on the render
result.}

Each grammar's layout function is extended to return a pair
\texttt{\{ scene: Scene, anchors: NodeAnchorRegistry \}} in addition to the
Scene.  The registry maps each node id to its bounding box and optional
connection ports, expressed in the cell's \textbf{local coordinate space}
(i.e., before composition-level translation and scaling).  Grammars that opt
in populate the registry from data they already hold during layout (the flow
grammar's \texttt{PlacedNode} is a direct source); grammars that have not yet
opted in return an empty registry, and links referencing their nodes degrade
gracefully.

\begin{lstlisting}[language=javascript,caption={NodeAnchorRegistry --- type specification (pseudo-TypeScript)}]
/** Port positions on the four cardinal sides of a node's bounding box. */
type CardinalSide = 'N' | 'S' | 'E' | 'W';

/**
 * Anchor for one node: bounding-box in local cell coordinates,
 * plus optional explicit port attachment points.
 * If ports is absent, the link layer derives port midpoints from
 * (x,y,w,h) at rendering time.
 */
interface NodeAnchor {
  x: number;                                   // bounding-box top-left x
  y: number;                                   // bounding-box top-left y
  w: number;                                   // bounding-box width
  h: number;                                   // bounding-box height
  ports?: Partial<Record<CardinalSide, { x: number; y: number }>>;
}

/** Maps diagram-local node id -> anchor in local cell coordinates. */
type NodeAnchorRegistry = Record<string, NodeAnchor>;

/**
 * Extended render result returned by linkable grammars.
 * Non-linkable grammars continue to return Scene only; the composition
 * layer treats an absent registry as {}.
 */
interface RenderResult {
  scene: Scene;
  anchors: NodeAnchorRegistry;   // empty ({}) for non-linkable grammars
}
\end{lstlisting}

\textit{Pros:} Explicit, typed contract authored by the grammar that knows its
node positions best; no scene walking; carries port information; each grammar
opts in independently; composable independently of the scene.

\medskip
\textit{Cons:} Requires changing grammar function return types; migration is
incremental (one grammar at a time).

\subsubsection{Recommended Implementation: Sidecar Anchors (Candidate b)}

\textbf{The recommended approach is the sidecar} \textbf{\texttt{NodeAnchorRegistry}} \textbf{(candidate b).}  The rationale:

\begin{enumerate}
  \item The grammar layout engine already owns the exact node geometry.  For
        the flow grammar, \texttt{PlacedNode.x}, \texttt{.y}, \texttt{.w},
        \texttt{.h} and the port coordinates \texttt{.rx}/\texttt{.lx}/\texttt{.cx}/\texttt{.by}
        are computed during Phase 4 (coordinate assignment) and available
        before scene emission.  Populating the registry is a two-line loop over
        the placed-node map --- no new computation required.
  \item The composition layer already tracks each cell's
        \texttt{(dx, dy, scale)} from the \texttt{translateAndScale} call in
        \texttt{layoutComposition} (see \texttt{packages/core/src/composition/layout.ts}).
        Transforming local anchors to poster coordinates is exact and trivial:
        \[
          x_{\mathrm{poster}} = x_{\mathrm{local}} \cdot s + d_x, \quad
          y_{\mathrm{poster}} = y_{\mathrm{local}} \cdot s + d_y, \quad
          w_{\mathrm{poster}} = w_{\mathrm{local}} \cdot s, \quad
          h_{\mathrm{poster}} = h_{\mathrm{local}} \cdot s
        \]
        where $s$ is the uniform scale factor and $(d_x, d_y)$ is the
        translation offset already computed by the grid layout pass.
  \item Candidate (a) would require reconstructing the same geometry from the
        scene tree, at the cost of an O($n$)-walk on the flattened primitive
        list, after the coordinate transform has already been applied.
  \item The opt-in migration path means zero risk to existing grammars:
        a grammar that has not yet emitted anchors returns \texttt{\{\}};
        links referencing it degrade to a WARN and are skipped
        (Section~\ref{sec:cdl-semantics}).
\end{enumerate}

\paragraph{Transformation to poster coordinates.}

After each cell's \texttt{RenderResult} is produced, the composition layer
performs the anchor transformation:

\begin{lstlisting}[language=javascript,caption={Anchor transformation to poster coordinates (pseudo-code)}]
function transformAnchors(
  registry: NodeAnchorRegistry,
  dx: number,    // cell's x offset in poster (from translateAndScale)
  dy: number,    // cell's y offset in poster
  scale: number, // uniform scale factor applied to the sub-scene
): NodeAnchorRegistry {
  const out: NodeAnchorRegistry = {};
  for (const [id, anchor] of Object.entries(registry)) {
    const pa: NodeAnchor = {
      x: anchor.x * scale + dx,
      y: anchor.y * scale + dy,
      w: anchor.w * scale,
      h: anchor.h * scale,
    };
    if (anchor.ports) {
      pa.ports = {};
      for (const [side, pt] of Object.entries(anchor.ports)) {
        pa.ports[side as CardinalSide] = {
          x: pt.x * scale + dx,
          y: pt.y * scale + dy,
        };
      }
    }
    out[id] = pa;
  }
  return out;
}
\end{lstlisting}

The resulting poster-coordinate registries from all cells are stored alongside
the placed-cell rectangles for use by the overlay routing pass.

\paragraph{Reusability argument.}

The node-anchor registry is valuable well beyond cross-diagram linking and
should be built regardless of whether the linking feature ships in the first
increment:

\begin{itemize}
  \item \textbf{Hit-testing:} An interactive SVG export can route mouse events
        to the correct diagram node by name, enabling click/hover handlers.
  \item \textbf{Tooltips:} Structured hover labels (from the Domain IR's node
        metadata) can be attached to the exact node bounding box.
  \item \textbf{Agent node-addressing:} The agent integration surface
        (Section~\ref{sec:agent-integration}) currently addresses diagrams by
        their IR structure.  A published anchor registry lets an agent
        reference ``cell A1, node \texttt{pay}'' by logical id rather than by
        pixel coordinates --- a stable, round-trip-safe reference.
  \item \textbf{Accessibility:} ARIA \texttt{role="img"} subtrees can carry
        \texttt{aria-label} attributes derived from each node's anchor id and
        IR label, improving screen-reader navigation of complex diagrams.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Overlay Routing}
\label{sec:cdl-overlay-routing}
% ─────────────────────────────────────────────────────────────────────────────

\paragraph{Overlay layer.}

Cross-diagram edges are rendered as a synthetic pass executed \emph{after} all
cell sub-scenes have been embedded.  The overlay primitives (line/path) are
appended to the poster's primitive list last, placing them above all cell
content in painter's-algorithm order.  No new Scene IR primitive type is
required: overlay edges are \texttt{PathPrimitive} (directed, with arrowhead)
or \texttt{LinePrimitive} (plain) constructed from the same kernel as
intra-diagram edges.

\paragraph{Port selection.}

Given source anchor $A$ (with cardinal ports $A_N$, $A_S$, $A_E$, $A_W$) and
target anchor $B$, the link layer selects the port pair $(p_A, p_B)$ that
minimises the Euclidean distance $\|p_A - p_B\|$.  Tie-breaking order: E
before W before S before N (favouring left-to-right reading direction).  When
explicit ports are absent from the registry, the four cardinal midpoints are
derived from the bounding box:
\[
  A_N = (x + w/2,\; y), \quad
  A_S = (x + w/2,\; y+h), \quad
  A_E = (x+w,\; y+h/2), \quad
  A_W = (x,\; y+h/2).
\]

\paragraph{Phase 1 routing: straight and single-elbow.}

The initial implementation uses two routing strategies selected by the
relative cell positions:

\begin{description}
  \item[Same row, adjacent columns] Straight horizontal segment from $p_A$
        to $p_B$, passing through the inter-cell gap.  This is the most common
        case (left-to-right flow) and produces clean, readable results with no
        routing logic.

  \item[Different rows or non-adjacent columns] Single-elbow (L-shaped) route:
        a horizontal segment from $p_A$ to a midpoint column, then a vertical
        segment to $p_B$'s row, then a horizontal segment to $p_B$.  The
        elbow column is placed at the midpoint between the two cell centres.
        All segments run through the inter-cell gap space and do not overlap
        cell bodies for standard grid layouts.
\end{description}

\paragraph{Phase 2 hardening: orthogonal routing.}

For complex poster layouts where the straight/elbow route would cross a cell
body, the composition engine can activate orthogonal routing:

\begin{enumerate}
  \item Build a set of \emph{forbidden rectangles} from each placed cell's
        bounding box, expanded outward by the theme's \texttt{gap} / 2.
  \item Compute the rectilinear shortest path from $p_A$ to $p_B$ that avoids
        all forbidden rectangles, using the inter-cell gap channels as routing
        channels.
  \item If no route exists (e.g., a cell is surrounded on all sides), emit a
        WARNING and fall back to the Phase 1 route drawn on top of the
        obstructing cell (readable but visually overlapping).
\end{enumerate}

Phase 2 routing is a \emph{hardening step}; Phase 1 covers the vast majority
of practical poster layouts and is the only mode required for the first
increment.

\paragraph{Dimension guard.}

The overlay routing pass applies dimension guards consistent with the
project's overall guard mindset
(Section~\ref{sec:composition-edge-cases}):

\begin{itemize}
  \item If the route must cross more than three cell bodies (unreachable via
        gap channels), emit \texttt{WARNING: link \textit{A.nodeId}
        $\to$ \textit{B.nodeId}: route crosses N cell bodies; overlay may be
        unreadable.}  Still render best-effort.
  \item If both endpoints resolve to the same cell, emit
        \texttt{WARNING: intra-cell link \textit{A.nodeId} $\to$
        \textit{A.nodeId2}: use an intra-diagram edge instead.}  The link is
        skipped (a cross-diagram link within one cell is semantically
        incoherent).
  \item If the computed route length (sum of segment lengths) exceeds three
        times the poster canvas diagonal, emit a WARNING.  This guards against
        degenerate posters where two cells are placed at extreme corners of a
        very large canvas.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Linkable-Type Rules}
\label{sec:cdl-linkable-types}
% ─────────────────────────────────────────────────────────────────────────────

A diagram type is \textbf{linkable} if and only if its layout engine can
produce a non-empty \texttt{NodeAnchorRegistry} --- that is, it authors stable
ids for its visual nodes and can map those ids to bounding boxes.

\begin{table}[h]
\centering\small
\caption{Linkable diagram types (v1)}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Grammar} & \textbf{Addressable unit} & \textbf{Node id source} \\
\midrule
\texttt{flowchart} / \texttt{graph} & Node & Author-assigned id (e.g.\ \texttt{A}, \texttt{pay}) \\
\texttt{classDiagram} & Class & Class name as written \\
\texttt{stateDiagram-v2} & State & State name / explicit id \\
\texttt{erDiagram} & Entity & Entity name as written \\
\texttt{C4Context} / \texttt{C4Container} / \texttt{C4Component} & Person, System, Container, Component & Alias string \\
\texttt{block-beta} & Block & Block id \\
\texttt{architecture-beta} & Service, Group & Id as written \\
\texttt{mindmap} & Node & Label-derived kebab-slug \\
\texttt{gitGraph} & Commit, Branch & Commit id or branch name \\
\bottomrule
\end{tabularx}
\label{tab:cdl-linkable}
\end{table}

\begin{table}[h]
\centering\small
\caption{Non-linkable diagram types (v1) --- excluded or deferred}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Grammar} & \textbf{Reason excluded} & \textbf{Degradation} \\
\midrule
\texttt{pie} & No stable node ids; data is label+value pairs & WARN + skip link \\
\texttt{sankey} & Edge-based flow; no persistent node ids & WARN + skip link \\
\texttt{quadrantChart} & Continuous 2-D axes; data points have no ids & WARN + skip link \\
\texttt{radar} & Series labels, not structural nodes & WARN + skip link \\
\texttt{packet-beta} & Byte-field labels, not structural nodes & WARN + skip link \\
\texttt{xychart-beta} & Continuous axis; no node abstraction & WARN + skip link \\
\texttt{timeline} & Events are date-keyed, not node-keyed & WARN + skip link \\
\texttt{sequenceDiagram} & Participants are stable identifiers; deferred to v2 & WARN + skip link \\
\texttt{gantt} & Bars are task-keyed; deferred to v2 & WARN + skip link \\
\bottomrule
\end{tabularx}
\label{tab:cdl-non-linkable}
\end{table}

\paragraph{Rule statement.}

\begin{quote}
A \texttt{link} endpoint is resolvable if and only if (1) the referenced cell
address is valid, (2) the cell's diagram type appears in
Table~\ref{tab:cdl-linkable}, and (3) the specified \texttt{nodeId} is present
in that cell's \texttt{NodeAnchorRegistry} after compilation.  Any endpoint
that fails any condition is \textbf{unresolvable}: the compiler emits one
\texttt{WARNING} per unresolvable endpoint, and the entire \texttt{link}
statement is omitted from the overlay.  A link with one valid endpoint and one
unresolvable endpoint is wholly omitted (a dangling edge has no sensible
routing target).
\end{quote}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Semantics and Degradation Contract}
\label{sec:cdl-semantics}
% ─────────────────────────────────────────────────────────────────────────────

\begin{quote}
\textbf{Cross-diagram link contract.}
Cross-diagram links are \emph{presentation-layer assertions on the
composition}.  They do not modify, extend, or reference any individual
diagram's Domain IR or Scene IR; each cell compiles as if \texttt{link}
statements did not exist.  Cross-diagram links are \emph{superset-only}: a
standard Mermaid renderer will reject a poster document containing
\texttt{link} statements (the keyword is outside Mermaid's grammar), which is
the intended portability signal.  All link resolution --- cell lookup, node
anchor lookup, port selection, and route computation --- occurs at composition
time, after every cell has compiled independently.  An unresolvable link
(unknown cell address, unknown node id within a valid cell, non-linkable
diagram type, or empty anchor registry) emits \textbf{exactly one
\texttt{WARNING} per unresolvable endpoint} and the link is \emph{silently
omitted}.  The poster renders in full; all other valid links render normally.
\textbf{Cross-diagram links are never fatal.}
\end{quote}

\paragraph{Mermaid portability.}

As with the \texttt{poster} keyword itself (Section~\ref{sec:superset-extensions}),
a file containing \texttt{link} statements cannot be rendered by a standard
Mermaid renderer.  This is intentional and clearly signals the superset
requirement.  Individual cell bodies (the diagram source embedded in each
\texttt{cell} block) remain standard Mermaid syntax and can be extracted and
rendered independently by any Mermaid renderer.

\paragraph{IR integrity.}

Because cross-diagram links are resolved entirely at composition time and
rendered as overlay primitives, they require no changes to any grammar's
Domain IR schema, to the Scene IR primitive types, or to the determinism
contract.  The \texttt{sceneHash} of any individual cell's sub-scene is
byte-identical to what it would be without the \texttt{link} statement.

\paragraph{Agent addressing.}

The \texttt{nodeId} token in a \texttt{link} statement is the same identifier
an agent would use when addressing a node via the structured IR path
(Section~\ref{sec:agent-integration}).  This alignment is intentional: the
anchor registry (Section~\ref{sec:cdl-anchor-registry}) is the shared
mechanism that makes both the cross-diagram link syntax and agent
node-addressing stable and consistent.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Engineering Prerequisites}
\label{sec:cdl-prereqs}
% ─────────────────────────────────────────────────────────────────────────────

The following work items must be completed before any cross-diagram link can
be drawn.  They are listed in dependency order.

\begin{enumerate}
  \item \textbf{Define \texttt{NodeAnchorRegistry} and \texttt{RenderResult}
        types} in \texttt{packages/core/src/scene.ts} (or a new
        \texttt{scene-anchors.ts}).  This is a pure type addition; no
        behaviour changes.

  \item \textbf{Extend \texttt{buildFlowScene}} to return
        \texttt{RenderResult}.  The flow grammar already computes
        \texttt{PlacedNode} records; emitting the registry requires only
        a loop over the placed-node map at the end of Phase 4.  This is
        the \emph{reference implementation} that verifies the approach
        end-to-end.

  \item \textbf{Extend linkable grammar layout functions} (one per grammar,
        per Table~\ref{tab:cdl-linkable}) to emit anchors.  Each extension is
        independent and can be shipped incrementally.  Non-extended grammars
        return \texttt{anchors: \{\}}.

  \item \textbf{Extend the composition layer} (\texttt{layoutComposition} in
        \texttt{packages/core/src/composition/layout.ts}) to (a) collect anchor
        registries from each cell's \texttt{RenderResult}, (b) apply the
        coordinate transform, and (c) build the poster-coordinate anchor map.

  \item \textbf{Implement the link parser} in the poster DSL
        (\texttt{packages/core/src/frontend/mermaid/poster.ts}): recognise
        \texttt{link} statements and store them alongside the \texttt{cells}
        array in \texttt{PosterDocument}.

  \item \textbf{Implement the overlay routing pass} after the cell embed loop
        in \texttt{layoutComposition}: resolve endpoints, select ports, compute
        routes, and append overlay primitives.

  \item \textbf{Implement graceful degradation}: WARN on unresolvable
        endpoints, skip the link, continue.

  \item \textbf{Extend the poster parser} to recognise \texttt{trace}
        statements (Section~\ref{sec:cdl-trace-syntax}) and populate
        \texttt{PosterDocument.traces: TraceRecord[]}.  Desugar each
        \texttt{TraceRecord} to the ordered sequence of atomic link pairs it
        implies, storing them as \texttt{link} entries tagged with their
        parent trace id.  This is a pure addition to the poster parser;
        the link-resolution, routing, and overlay machinery from items 5--7
        are reused without modification.

  \item \textbf{Assign categorical-palette colours} to \texttt{TraceRecord}s
        at composition time: iterate traces in declaration order and assign
        \texttt{categorical[i \% n]} from the active theme's data palette
        (Section~\ref{sec:contract-palette}).  Store the resolved colour in
        \texttt{TraceRecord.color} for use by the overlay rendering pass.

  \item \textbf{Emit the trace legend}: after all overlay edges are appended,
        generate a legend block below the poster canvas listing each named
        trace with its colour swatch and type pill.  Suppress if
        \texttt{trace legend: off} is set in the poster frontmatter.
\end{enumerate}

Items 1, 5, and 8 are pure additions.  Items 2--4 and 6--7 are incremental
extensions to existing functions with no breaking changes to current callers
(the \texttt{Scene} return type is replaced by \texttt{RenderResult} for
linkable grammars; callers that ignore the \texttt{anchors} field are
unaffected).  Items 9--10 extend only the composition layer and overlay pass.

% =============================================================================
% §30b.8 — The trace Abstraction: Named, Typed, Ordered Multi-Hop Paths
% =============================================================================
\subsection{The \texttt{trace} Abstraction: Named, Typed, Ordered Paths}
\label{sec:cdl-trace}

\begin{quote}
\textbf{Headline use case.}  A poster assembled from layered system views ---
a C4 Context diagram, a Container diagram, a Component diagram, an ER schema,
a test suite --- is a natural representation of \emph{system architecture
across abstraction levels}.  A single user story, requirement, or runtime
request touches nodes in several of those layers simultaneously.  The
\texttt{trace} construct names, types, and orders that multi-layer thread,
turning a poster from a side-by-side collection of views into a first-class
\textbf{system traceability artefact}.
\end{quote}

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Concept}
\label{sec:cdl-trace-concept}

A \textbf{trace} is a \emph{named, ordered, optionally-typed multi-hop path}
of cross-diagram links.  Where a single \texttt{link} is one atomic directed
edge between two nodes in different cells, a \texttt{trace} is a path
$A \to B \to C \to \cdots$ stitching together a logical thread that runs
across system layers or diagram views in a fixed order.

Three properties distinguish a trace from a loose collection of links:

\begin{description}
  \item[Named.]  A trace carries a human-readable name (a quoted string) that
    identifies the logical thread it represents --- for example,
    \texttt{"Login flow"} or \texttt{"REQ-12"}.  The name is the stable
    handle used for highlight, filter, and legend display.

  \item[Ordered.]  The hops of a trace form a sequence, not a bag of edges.
    Ordering encodes direction and traversal: from origin (e.g.\ a requirement
    or user action) through intermediate layers (components, services) to
    terminus (e.g.\ a test, a database table, or a deployed service).
    Ordering is what makes a trace a \emph{path}, not a cluster.

  \item[Typed.]  An optional type token drawn from a shared vocabulary
    (Section~\ref{sec:cdl-trace-vocab}) annotates every hop in the trace with
    the same semantic relationship kind.  Typing makes traceability uniform:
    a \texttt{satisfies} trace across a poster carries exactly the same
    meaning as a \texttt{satisfies} edge inside a \texttt{requirementDiagram}
    cell.
\end{description}

\paragraph{Relation to \texttt{link}: desugaring.}

The \texttt{link} statement remains the \textbf{primitive}.  A \texttt{trace}
\emph{desugars} to an ordered sequence of atomic \texttt{link}s \emph{plus}
a \texttt{TraceRecord} (name, type, ordered member link list).  Specifically,
a trace with $n$ hops $h_1 \to h_2 \to \cdots \to h_n$ lowers at composition
time to the $n - 1$ atomic links $\{h_1 \to h_2,\; h_2 \to h_3,\; \ldots,\;
h_{n-1} \to h_n\}$ together with a \texttt{TraceRecord} that groups those
links under the trace name, type, and ordinal index.  The link-resolution,
port-selection, and overlay-routing machinery from Sections
\ref{sec:cdl-anchor-registry}--\ref{sec:cdl-semantics} apply unchanged to
each desugared hop.

\begin{lstlisting}[language=,caption={Desugaring a trace to atomic links + group record}]
-- Input (poster DSL):
trace "Login flow" : A1.Login -> B1.AuthAPI -> C1.UserService -> D1.users

-- Desugars to three atomic links:
link A1.Login    --> B1.AuthAPI      -- hop 1->2  (trace member 0)
link B1.AuthAPI  --> C1.UserService  -- hop 2->3  (trace member 1)
link C1.UserService --> D1.users     -- hop 3->4  (trace member 2)
-- PLUS a TraceRecord:
-- { name: "Login flow", type: none,
--   members: [ link[0], link[1], link[2] ] }
\end{lstlisting}

A \texttt{link} statement that is not part of any \texttt{trace} continues to
behave exactly as specified in Section~\ref{sec:cdl-syntax}: a standalone
overlay edge with no trace-group affiliation and no categorical colour.

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Syntax}
\label{sec:cdl-trace-syntax}

The \texttt{trace} statement is a poster-level declaration written alongside
\texttt{cell} and \texttt{link} definitions.  Its abstract grammar is:

\begin{lstlisting}[language=,caption={Trace statement --- abstract grammar}]
trace "<name>" [<type>] : <hop> <arrow> <hop> { <arrow> <hop> }

<name>  ::= quoted string (UTF-8; any character except unescaped `"`)

<type>  ::= (* optional; drawn from the trace-type vocabulary,
               Section~\ref{sec:cdl-trace-vocab} *)
            "satisfies" | "derives"  | "verifies" | "refines"
          | "traces"    | "contains" | "copies"
          | "calls"     | "flowsTo"  | "mapsTo"

<hop>   ::= <cellAddr> "." <nodeId>
            (* cellAddr and nodeId same as link, Section~\ref{sec:cdl-syntax} *)

<arrow> ::= "-->"   (* solid directed --- default for most types *)
          | "-..->" (* dashed directed --- for weak or inferred links *)
\end{lstlisting}

Two concrete examples:

\begin{lstlisting}[language=,caption={Trace syntax: untyped presentation-flow and typed requirement trace}]
-- Untyped trace: a named presentation-flow path across four cells
trace "Login flow" : A1.Login -> B1.AuthAPI -> C1.UserService -> D1.users

-- Typed trace: requirement traceability using the requirement vocabulary
trace "REQ-12" satisfies : A1.Req12 --> B1.AuthService --> C1.AuthTest
\end{lstlisting}

\paragraph{Arrow forms in traces.}

Inside a \texttt{trace} path the arrow token determines the edge style of
every desugared link in that trace.  When the type token is present the
compiler selects the canonical arrow form for that type per
Table~\ref{tab:cdl-trace-types}; the arrow token in the source may be omitted
when a type is supplied.  When omitted entirely the default is \texttt{-->}.

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Trace-Type Vocabulary}
\label{sec:cdl-trace-vocab}

The trace type is drawn from two groups.  The first group is the
\textbf{requirement-relationship vocabulary} used by the
\texttt{requirementDiagram} grammar, grounded in
\texttt{packages/core/src/grammars/requirement/types.ts}
(\texttt{RequirementRelKind}).  The Zod schema in
\texttt{packages/core/src/grammars/requirement/schema.ts} enumerates
the identical seven values.  The second group is a small set of
\textbf{presentation-flow types} for architectural and runtime diagrams.

Reusing the requirement-grammar vocabulary ensures that a traceability
assertion is semantically identical whether expressed as a
\texttt{requirementDiagram} edge inside one cell or as a typed \texttt{trace}
across poster cells.  An author can graduate a single-cell requirement diagram
into a cross-view poster trace without changing the relationship kind.  Tools
that index or query the composition IR see uniform type tokens in both
contexts.

\begin{table}[h]
\centering\small
\caption{Trace-type vocabulary and default presentation}
\begin{tabularx}{\textwidth}{p{2.2cm}p{2.2cm}p{1.6cm}p{2.2cm}X}
\toprule
\textbf{Type token} & \textbf{Group} & \textbf{Default arrow} & \textbf{IR source} & \textbf{Semantics} \\
\midrule
\texttt{satisfies}  & Requirement & \texttt{-{}-{}>}  & \texttt{RequirementRelKind} & Component satisfies a requirement \\
\texttt{derives}    & Requirement & \texttt{-.->} & \texttt{RequirementRelKind} & Requirement derived from another \\
\texttt{verifies}   & Requirement & \texttt{-{}-{}>}  & \texttt{RequirementRelKind} & Test verifies a requirement \\
\texttt{refines}    & Requirement & \texttt{-.->} & \texttt{RequirementRelKind} & Requirement refines a higher-level one \\
\texttt{traces}     & Requirement & \texttt{-.->} & \texttt{RequirementRelKind} & Generic traceability link \\
\texttt{contains}   & Requirement & \texttt{-{}-{}>}  & \texttt{RequirementRelKind} & Requirement contains a sub-requirement \\
\texttt{copies}     & Requirement & \texttt{-.->} & \texttt{RequirementRelKind} & Requirement copied from another \\
\midrule
\texttt{calls}      & Presentation & \texttt{-{}-{}>}  & (new; poster-layer) & A service or component calls another \\
\texttt{flowsTo}    & Presentation & \texttt{-{}-{}>}  & (new; poster-layer) & Data or control flows to a node \\
\texttt{mapsTo}     & Presentation & \texttt{-.->} & (new; poster-layer) & A model element maps to a view element \\
\bottomrule
\end{tabularx}
\label{tab:cdl-trace-types}
\end{table}

When \texttt{<type>} is omitted, the trace is \textbf{untyped}: all desugared
links receive the plain directed arrow (\texttt{-->}) and no semantic label or
pill is rendered.  Untyped traces are valid and useful for pure
presentation-flow paths where no traceability semantics are implied.

The three presentation-flow types (\texttt{calls}, \texttt{flowsTo},
\texttt{mapsTo}) are poster-layer additions with no counterpart in the
\texttt{requirementDiagram} grammar.  They are excluded from
\texttt{RequirementRelKind} deliberately: they carry no formal requirements-
engineering semantics and their home is the composition layer, not the domain
grammar.

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Rationale: Why Traces Beat Loose Links}
\label{sec:cdl-trace-rationale}

Loose individual \texttt{link} statements can express the same set of edges as
a trace, but without the structure that makes those edges useful as a
traceability artefact.  Three properties of traces over loose links:

\begin{description}
  \item[Named and highlightable.]
    A trace can be emphasised \emph{as a unit}.  Hovering or selecting
    \texttt{"Login flow"} in the legend highlights every hop in that path
    simultaneously, dimming all other overlay edges and uninvolved cells.
    Filtering to a single trace hides all other links.  A loose collection of
    links with the same topology has no identity, no unit of selection, and no
    filter target.

  \item[Typed and consistent.]
    A typed trace carries traceability semantics (e.g.\ \texttt{satisfies},
    \texttt{verifies}) that are uniform across the poster because they come
    from the same \texttt{RequirementRelKind} vocabulary used inside
    \texttt{requirementDiagram} cells.  A reader of a poster trace and a
    reader of a requirement-diagram edge see the same semantic label.  Tools
    that consume the composition IR can index traceability uniformly regardless
    of which representation was used.

  \item[Ordered and directional.]
    A path $A \to B \to C$ encodes traversal direction and origin; a bag of
    edges $\{A \to B,\; B \to C\}$ does not, in isolation, identify that the
    path begins at $A$ and terminates at $C$.  For traceability, direction
    matters: \emph{requirement REQ-12 is satisfied by AuthService which is
    verified by AuthTest} is different from the reverse, and the ordered trace
    makes that unambiguous.
\end{description}

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Presentation Semantics}
\label{sec:cdl-trace-presentation}

\paragraph{Per-trace colour from the data palette.}

Each named trace is assigned a distinct colour drawn from the theme's
\textbf{categorical data palette} (Section~\ref{sec:contract-palette}).  This
is the same palette used by chart-family components (bar charts, XY charts,
pie charts) to assign per-series colours.  Reusing the categorical palette
ensures that trace colours are:

\begin{enumerate}
  \item \textbf{Perceptually distinct} from one another --- the categorical
    palette is designed for discrimination across 6--12 concurrent categories.

  \item \textbf{On-brand and theme-coherent} --- the palette is defined by the
    active contract theme, so \texttt{executive}, \texttt{pastel},
    \texttt{terminal}, and all other themes provide trace-appropriate
    categorical colours automatically and consistently with every other diagram
    on the poster.

  \item \textbf{Non-clashing with diagram-internal colours} --- diagram
    components consume the \emph{semantic role palette} (surface, ink, accent,
    muted, border), while traces consume the \emph{data palette}.  The two
    sub-palettes are designed within each theme never to collide
    (Section~\ref{sec:contract-palette}).
\end{enumerate}

Traces are assigned colours by declaration order: the first \texttt{trace}
statement receives \texttt{categorical[0]}, the second \texttt{categorical[1]},
and so on.  If the number of traces exceeds the categorical palette length $n$,
the palette wraps modulo $n$ with a 20\% lightness offset applied on each
cycle to maintain visual distinction.

\paragraph{Trace legend.}

A \textbf{trace legend} is automatically generated and appended at the bottom
of the poster canvas when at least one named trace is present.  Each legend
entry shows:

\begin{itemize}
  \item A colour swatch (small filled rectangle) in the trace's assigned
    categorical colour.
  \item The trace name (the quoted string as authored).
  \item The type token, if present, rendered as a \texttt{«type»} pill styled
    by the theme's \texttt{overlayLabelFont} token --- the same pill style
    used for \texttt{requirementDiagram} relationship labels.
\end{itemize}

The legend can be suppressed with a \texttt{trace legend: off} directive in
the poster frontmatter.  When no named traces exist, no legend is emitted.

\paragraph{Highlight, dim, and filter modes.}

Traces introduce three presentation-layer interaction modes at the composition
level.  These are specified here at the design level; rendering detail
(interactive SVG event handlers, static pre-rendered variants, CLI
\texttt{--active-trace} flags) is deferred to the backend implementation.

\begin{description}
  \item[\textbf{Highlight mode} (one trace active).]  When a trace is
    activated (e.g.\ hover on a legend entry or programmatic selection via the
    agent API): all edges belonging to the active trace render at full opacity
    with a luminosity boost of the trace colour; all other overlay edges render
    at 30\% opacity; cells containing no hop of the active trace render at 40\%
    opacity; cells containing at least one hop render at full opacity.

  \item[\textbf{Dim mode} (any edge hovered, no specific trace selected).]
    All overlay edges render at full opacity.  Bare \texttt{link} statements
    (not part of any trace) render at 60\% opacity.  No cells are dimmed.

  \item[\textbf{Filter mode} (trace selection).]  Only selected trace(s) are
    visible; all other overlay edges and non-participating cells are hidden.
    Filter mode is toggled via the legend (click a legend entry) or
    programmatically via the agent API
    (Section~\ref{sec:cdl-trace-agent}).  Multiple traces may be simultaneously
    active in filter mode.
\end{description}

\paragraph{Degradation contract for traces.}

Pathological cases in trace resolution follow the same ``warn and render
the resolvable portion'' principle as atomic links, with refinements specific
to the multi-hop structure:

\begin{description}
  \item[\textbf{One unresolvable hop} ($h_k$ fails).]  The compiler emits
    \texttt{WARNING: trace "\textit{name}" hop \textit{k}: \textit{<reason>}
    --- hop omitted.}  The trace renders as two sub-paths: $h_1 \to \cdots
    \to h_{k-1}$ and $h_{k+1} \to \cdots \to h_n$, both in the trace's
    assigned colour.  The trace name still appears in the legend.

  \item[\textbf{All hops unresolvable}.]  The trace is wholly omitted.
    The compiler emits \texttt{WARNING: trace "\textit{name}": no resolvable
    hops --- trace skipped.}  No legend entry is generated.

  \item[\textbf{Duplicate hop} (same node appears at positions $i$ and $j$).]
    The compiler emits \texttt{WARNING: trace "\textit{name}" hop \textit{j}:
    node \textit{X.Y} already appears at position \textit{i} --- cycle
    detected; loop arc rendered.}  A self-loop arc is drawn at the repeated
    node.

  \item[\textbf{Fewer than two resolvable hops}.]  A trace cannot form an
    edge without two resolved endpoints.  The compiler emits \texttt{WARNING:
    trace "\textit{name}": fewer than 2 resolvable hops --- trace skipped.}

  \item[\textbf{Hop routing crosses cell bodies}.]  Follows the atomic-link
    routing degradation (Section~\ref{sec:cdl-overlay-routing}): Phase~1
    fallback is used per individual desugared hop; an additional warning is
    emitted if the hop crosses more than three cell bodies.
\end{description}

All trace degradation is non-fatal.  The poster compiles and renders; only
the affected hop segments are omitted or replaced by fallback arcs.  Other
traces and all standalone \texttt{link} statements are unaffected.

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Worked Examples}
\label{sec:cdl-trace-examples}

% ---- Example 1: C4 architecture drill-down ----
\paragraph{Example 1 --- C4 architecture drill-down (context to container to component to data).}

The classic C4 model organises a software architecture across four nested
abstraction levels.  A four-cell poster places one C4 view per cell and
connects them with a single \texttt{trace} that represents one user journey
through the stack: from the context-level user persona, through a container,
down into a component, and into the data layer.

\begin{lstlisting}[language=,caption={C4 drill-down: one trace across four architecture abstraction levels}]
---
theme: executive
layout: grid 2x2
---
poster "E-Commerce Platform -- Login Journey"

  cell A1: C4Context
    Person(user, "Customer", "Web or mobile shopper")
    System(webapp, "Web Application", "React SPA")
    System_Ext(idp, "Identity Provider", "Auth0")
    Rel(user, webapp, "Uses", "HTTPS")
    Rel(webapp, idp, "Delegates auth", "OIDC")

  cell B1: C4Container
    Container(spa, "SPA", "React", "Browser client")
    Container(api, "API Gateway", "Node/Express", "REST")
    Container(authsvc, "Auth Service", "Go", "gRPC")
    Rel(spa, api, "Calls", "REST/HTTPS")
    Rel(api, authsvc, "Authenticates via", "gRPC")

  cell A2: C4Component
    Component(loginctrl, "LoginController", "React", "Login page")
    Component(authclient, "AuthClient", "TS", "OIDC adapter")
    Rel(loginctrl, authclient, "Invokes")

  cell B2: classDiagram
    class User
    class Session
    class Token
    User "1" --> "*" Session : owns
    Session --> Token : contains

trace "Login journey" calls :
  A1.user --> B1.spa --> B1.api --> B1.authsvc --> B2.Token
\end{lstlisting}

The \texttt{calls} trace renders in \texttt{categorical[0]} of the executive
theme.  A legend entry labelled ``Login journey / calls'' appears beneath the
poster grid.  The intra-cell sub-hop \texttt{B1.spa}\,$\to$\,\texttt{B1.api}
is within the same cell; the compiler emits a warning and renders a short
intra-cell loopback arc styled in the trace colour, then resumes the
cross-diagram routing for the next hop.

% ---- Example 2: Distributed request trace ----
\paragraph{Example 2 --- Distributed request trace across heterogeneous diagram types.}

A distributed system poster combines a flowchart (entry-point logic), a
sequence diagram (inter-service protocol, deferred/v2), an architecture
diagram (deployed topology), and an ER diagram (data schema) in one poster.
A single trace named for the operation represents one end-to-end request.

\begin{lstlisting}[language=,caption={Distributed request trace across four diagram types}]
---
theme: executive
layout: grid 2x2
---
poster "Payment Processing -- Charge Request"

  cell A1: flowchart LR
    entry[API Entry] --> validate[Validate Payload]
    validate --> auth[Authorise]
    auth --> charge[Charge]

  cell B1: sequenceDiagram
    APIGateway->>PaymentService: POST /charge
    PaymentService->>FraudEngine: check(payload)
    FraudEngine-->>PaymentService: score
    PaymentService->>Ledger: debit(account)

  cell A2: architecture-beta
    service APIGateway(server, "API Gateway", "us-east-1")
    service PaymentService(server, "Payment Svc", "us-east-1")
    service FraudEngine(server, "Fraud Engine", "eu-west-1")
    service Ledger(database, "Ledger DB", "us-east-1")
    APIGateway --> PaymentService
    PaymentService --> FraudEngine
    PaymentService --> Ledger

  cell B2: erDiagram
    ACCOUNT ||--o{ TRANSACTION : has
    TRANSACTION }|--|| CHARGE_EVENT : records
    CHARGE_EVENT }o--|| FRAUD_SCORE : scored_by

trace "Charge request" flowsTo :
  A1.entry --> A1.charge --> A2.PaymentService --> B2.TRANSACTION
\end{lstlisting}

The hop \texttt{A1.entry}\,$\to$\,\texttt{A1.charge} is intra-cell (both
in the flowchart); the compiler warns and renders a loopback arc, then
continues cross-diagram.  The sequence diagram cell \texttt{B1} is deliberately
not included in this trace because \texttt{sequenceDiagram} is non-linkable in
v1 (Table~\ref{tab:cdl-non-linkable}); it remains visible on the poster as
context.  The trace renders two cross-diagram hops:
\texttt{A1}\,$\to$\,\texttt{A2} (flowchart $\to$ architecture) and
\texttt{A2}\,$\to$\,\texttt{B2} (architecture $\to$ ER), both coloured in
\texttt{categorical[0]}.

% ---- Cell spans (poster DSL) ----
\subsubsection{Cell Spans in the Poster DSL}
\label{sec:cdl-cell-spans}

By default each \texttt{cell} header addresses a single grid cell.  To let a
single diagram occupy a rectangular block of the grid, the poster DSL accepts a
\emph{span} in the cell header.  Three equivalent forms are supported:

\begin{lstlisting}[language=,caption={Cell span forms --- all three place one cell in column~B spanning rows 1--2 (rowSpan~2)}]
cell B1:B2:    classDiagram   (* Excel range: top-left : bottom-right *)
cell [0,1]:[1,1]: classDiagram   (* bracket range, 0-indexed row,col *)
cell B1 rowspan 2: classDiagram   (* keyword form; colspan N also accepted *)
\end{lstlisting}

\noindent The range forms derive \texttt{colSpan}/\texttt{rowSpan} from the
inclusive distance between the two endpoints; the keyword form sets them
directly.  Spans feed straight into the composition engine's existing
\texttt{Cell.colSpan}/\texttt{Cell.rowSpan} fields, and the grid layout
distributes the spanning cell's intrinsic size across the tracks it covers so
that a tall hub cell aligns flush with the combined height of the cells beside
it.  Single-cell addressing is unchanged.

% ---- Example 3: Requirements traceability (typed, multi-trace) ----
\paragraph{Example 3 --- Requirements traceability with typed traces and a trace legend.}

This is the flagship use case for typed traces.  A requirement in one cell is
satisfied by a component in the hub cell, which is verified by a test in a
third cell.  Following the directive's hybrid layout, the two \emph{wide}
flowcharts (requirements and tests) are stacked in the left column and the
\emph{tall} class diagram spans both rows on the right as the hub.  Two
separate \texttt{satisfies} traces make two independent traceability chains
visible simultaneously, each in its own colour, with a generated legend.

\begin{lstlisting}[language=,caption={Multi-typed requirements traceability poster --- hybrid layout: wide flows stacked left, tall class hub spanning two rows on the right}]
---
theme: executive
layout: grid 2x2
---
poster "Authentication System — Requirements Traceability"

  cell A1: flowchart LR
    SPEC[Spec §4.2: Auth NFRs] --> REQ12[REQ-12: Authenticate < 500ms]
    SPEC --> REQ13[REQ-13: Token expires in 24h]

  cell A2: flowchart LR
    SUITE[CI auth-suite] --> AUTHT[auth_latency_test]
    SUITE --> TOKT[token_expiry_test]

  cell B1:B2: classDiagram
    class AuthController {
      +login(cred): Token
      +validate(tok): bool
    }
    class TokenService {
      +issue(uid): Token
      +revoke(tok): void
    }
    class SessionCache {
      +store(tok, ttl): void
      +lookup(tok): Session
    }
    AuthController --> TokenService : issues
    TokenService --> SessionCache : stores

trace "REQ-12" satisfies : A1.REQ12 --> B1.AuthController --> A2.AUTHT
trace "REQ-13" satisfies : A1.REQ13 --> B1.TokenService --> A2.TOKT
\end{lstlisting}

The cell header \texttt{cell B1:B2} uses the \emph{span} form of the poster DSL
(Section~\ref{sec:cdl-cell-spans}): the Excel range \texttt{B1:B2} places one
cell in column~B spanning rows~1--2, i.e. \texttt{rowSpan = 2}.  This is the
hybrid layout the directive calls for: the two \emph{wide} flowcharts (the
requirements in \texttt{A1} and the tests in \texttt{A2}) are stacked in the
left column, and the naturally \emph{tall} class diagram is placed to their
right as a single hub cell spanning their combined height.  Each
\texttt{satisfies} trace threads out of a requirement node, directly through the
central gutter into the hub class, then back out to the matching test node.

\ourdiagram[\linewidth]{trace-poster}{%
  Rendered hybrid poster: two \texttt{satisfies} traces
  (\textcolor[HTML]{1F497D}{REQ-12} and \textcolor[HTML]{2E86AB}{REQ-13})
  weave a traceability thread from the requirement nodes (top-left cell) and
  test nodes (bottom-left cell) through the tall service-class hub (right cell,
  spanning both rows).  Each trace routes directly through the central gutter
  --- lane-separated by colour, with arrowheads landing at cell boundaries and
  no segment crossing a non-endpoint node.  The auto-generated trace legend at
  the bottom shows each trace's colour swatch, name, and \texttt{«satisfies»}
  type pill --- rendered directly by the compiler from the listing above.}

The two traces render in \texttt{categorical[0]} and \texttt{categorical[1]}
of the executive theme respectively.  The auto-generated legend beneath the
poster grid reads:

\begin{center}\small
\begin{tabular}{lll}
\textbf{[swatch]} & REQ-12 & \textsf{«satisfies»} \\
\textbf{[swatch]} & REQ-13 & \textsf{«satisfies»} \\
\end{tabular}
\end{center}

A reviewer hovers ``REQ-12'' in the legend to highlight only the first
colour thread --- from the requirement cell through \texttt{AuthController} to
\texttt{auth\_latency\_test} --- while all other cells and the REQ-13 trace dim
to 40\%
opacity.  The filter mode lets a reviewer isolate a single requirement's
complete traceability chain on a poster that may contain dozens of requirements
and components.  This is the core interaction model for poster-scale
traceability review.

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Rendered Example: Two Typed Traces in the Hybrid Layout}
\label{sec:cdl-trace-rendered}

The following poster was rendered by the Timeline compiler (Phase~B,
\texttt{crosslink-poster.mmd}).  It adopts the same hybrid layout as the worked
example above: the wide requirement and test flowcharts are stacked in the left
column, and the tall service class diagram spans both rows as the hub on the
right.  Two named, typed \texttt{satisfies} traces thread through it.

\ourdiagram[0.97\linewidth]{crosslink-poster}{%
  Rendered hybrid poster: two \texttt{satisfies} traces (\textcolor[HTML]{1F497D}{REQ-01 chain}
  and \textcolor[HTML]{2E86AB}{REQ-02 chain}) weave a traceability thread from
  the requirement nodes (top-left cell) and test nodes (bottom-left cell)
  through the service class hub (right cell, spanning both rows).  Each trace
  routes directly through the central gutter, lane-separated by colour.  The
  trace legend at the bottom shows each trace's colour swatch, name, and
  \texttt{«satisfies»} type pill.}

\noindent The categorical data palette provides perceptually-distinct trace
colours that are on-brand with the \texttt{executive} theme and do not clash
with the diagram-internal navy border and ink colours.  The legend beneath the
grid makes the two traceability chains immediately identifiable at a glance.

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Agent Integration}
\label{sec:cdl-trace-agent}

Traces are first-class artefacts in the IR/composition API surface
(Section~\ref{sec:agent-integration}).  An agent does not need to manipulate
poster DSL text to assert or query traceability; it operates on the structured
\texttt{TraceRecord} type:

\begin{lstlisting}[language=javascript,caption={TraceRecord and TraceHop --- IR types for the agent API (pseudo-TypeScript)}]
/** A single hop in a trace: canonical cell address + node id. */
interface TraceHop {
  cell:   string;   // canonical Excel address ("A1", "B2", ...)
  nodeId: string;   // node identifier within that cell's diagram
}

/**
 * Trace-type vocabulary.
 * Requirement group: identical to RequirementRelKind in
 *   packages/core/src/grammars/requirement/types.ts
 * Presentation-flow group: poster-layer only.
 */
type TraceType =
  | 'satisfies' | 'derives'  | 'verifies'   // RequirementRelKind
  | 'refines'   | 'traces'   | 'contains'   // RequirementRelKind
  | 'copies'                                 // RequirementRelKind
  | 'calls'     | 'flowsTo'  | 'mapsTo';    // presentation-flow (poster-layer)

/** A named, typed, ordered multi-hop trace: the composition IR unit. */
interface TraceRecord {
  name:    string;      // human-readable identifier; must be unique per poster
  type?:   TraceType;   // omitted = untyped trace
  hops:    TraceHop[];  // ordered list of hops; length >= 2
  color?:  string;      // hex string; assigned at render time from categorical palette
}
\end{lstlisting}

An agent can use this API to:

\begin{itemize}
  \item \textbf{Assert traceability programmatically.}  Construct a
    \texttt{TraceRecord} with a name, type, and ordered hops and push it to
    \texttt{PosterDocument.traces}.  This is a structured traceability
    artefact that replaces ad-hoc comment annotations or manually maintained
    link lists.  The artefact is reproducible, diffable across poster
    revisions, and queryable without parsing diagram source.

  \item \textbf{Query traces by type.}  Filter
    \texttt{PosterDocument.traces} by \texttt{type} to enumerate all
    \texttt{satisfies} traces (requirement coverage) or all \texttt{verifies}
    traces (test coverage).

  \item \textbf{Activate filter mode.}  Set
    \texttt{PosterRenderOptions.activeTraces} to a list of trace names to
    render a filtered view with only the specified traces visible.  This lets
    an agent generate specialised traceability views without modifying the
    poster source.

  \item \textbf{Detect coverage gaps.}  Walk the \texttt{requirementDiagram}
    Domain IR in cell~$A$ to enumerate all requirements, then check whether
    each has a corresponding \texttt{satisfies} trace hop as its origin.  Flag
    requirements that have no such trace as uncovered --- a gap report that
    drives the next authoring action.
\end{itemize}

A \texttt{trace} is therefore not merely a drawing command: it is a
\textbf{structured, typed, agent-queryable traceability artefact}.  An
agent-generated traceability matrix is reproducible, queryable, and diffable
across poster revisions --- properties that no graphical or freeform annotation
scheme can provide.

% ─────────────────────────────────────────────────────────────────────────────
\subsection*{Cross-references}
% ─────────────────────────────────────────────────────────────────────────────

\begin{itemize}
  \item Composition / Posters (poster DSL, cell addressing):
        Section~\ref{sec:composition} and Section~\ref{sec:composition-posters}
  \item Dual cell addressing (bracket and Excel forms):
        Section~\ref{sec:composition-posters}
  \item Scene IR primitives (\texttt{GroupPrimitive.id},
        \texttt{PathPrimitive}): Section~\ref{sec:scene-ir} and
        \texttt{packages/core/src/scene.ts}
  \item \texttt{translateAndScale} coordinate helper:
        Section~\ref{sec:composition-embed} (paragraph \ref{para:translate-scale-helper})
  \item Superset extension mechanisms (\texttt{poster} keyword,
        superset-only features): Section~\ref{sec:superset-extensions}
  \item Agent integration surface (IR-as-agent-API, node-addressing):
        Section~\ref{sec:agent-integration}
  \item Dimension guard mindset: Section~\ref{sec:composition-edge-cases}
  \item Theme data palette / categorical sub-palette (trace colour source):
        Section~\ref{sec:contract-palette} and Section~\ref{sec:themes}
  \item Requirement-relationship vocabulary (\texttt{RequirementRelKind},
        seven values): \texttt{packages/core/src/grammars/requirement/types.ts}
        and \texttt{packages/core/src/grammars/requirement/schema.ts}
\end{itemize}
