\section{Packaging \& Repository Strategy}
\label{sec:packaging}

This section defines the packaging strategy for the diagram compiler: how modules are organized in the repository, how they are published, and the incremental path from the current timeline-only codebase to the full grammar family.

\subsection{Strategic Principle}

The diagram compiler lives \textbf{beside timeline in the same monorepo on a shared kernel}---not inside timeline's IR (which would bloat it), not in a separate repo (which would duplicate the engine and drift).

\begin{figure}[h]
\centering
\begin{Verbatim}[frame=single,fontsize=\small]
packages/
  kernel/          @diagram-compiler/kernel
  timeline/        @diagram-compiler/timeline
  flow/            @diagram-compiler/flow        (future)
  graph/           @diagram-compiler/graph       (future)
  comparison/      @diagram-compiler/comparison  (future)
  composition/     @diagram-compiler/composition (future)
  cli/             @diagram-compiler/cli
  mcp/             @diagram-compiler/mcp
\end{Verbatim}
\caption{Target package structure}
\label{fig:package-structure}
\end{figure}

\subsection{Compounding Payoff}

This structure creates compounding value:

\begin{itemize}
  \item \textbf{Animation} added to the kernel is inherited by all grammars
  \item \textbf{New backends} serve all grammars immediately
  \item \textbf{Theme improvements} apply everywhere
  \item \textbf{Determinism tests} cover all grammars
  \item \textbf{Icon additions} are available to all grammars
  \item \textbf{Security patches} to the kernel propagate to all grammars
\end{itemize}

A new grammar starts with the full feature set: determinism, theming, SVG/PNG/PDF/PPTX output, animation support, validation---all for free.

\subsection{The Incremental Path}

The current codebase (\texttt{packages/core}) is timeline-only. Extracting the kernel requires care to avoid a premature big-bang refactor. The recommended path:

\subsubsection{Phase 0: Draw the Seam (Current)}

\textbf{Goal:} Establish the kernel/grammar boundary \emph{inside} \texttt{packages/core}, without restructuring packages.

\textbf{Actions:}
\begin{enumerate}
  \item Create \texttt{src/kernel/} directory within \texttt{packages/core}
  \item Move kernel components (Scene IR, backends, themes, icons, font metrics) into \texttt{src/kernel/}
  \item Move timeline-specific components (Timeline IR, layout engine) into \texttt{src/timeline/}
  \item Add lint rule: \texttt{src/kernel/} may not import from \texttt{src/timeline/}
  \item Verify all tests pass
\end{enumerate}

\textbf{Outcome:} The kernel/timeline boundary is enforced by convention and linting, but no packages are split. The monorepo structure is unchanged.

\subsubsection{Phase 1: Prove Grammar-Agnosticism}

\textbf{Goal:} Build the first new grammar (Flow) as a kernel-only consumer, proving the architecture.

\textbf{Actions:}
\begin{enumerate}
  \item Create \texttt{src/flow/} within \texttt{packages/core}
  \item Implement Flow grammar (IR, validation, layout) using only \texttt{src/kernel/} imports
  \item Add Flow grammar tests
  \item Verify the kernel imports nothing from \texttt{src/timeline/} or \texttt{src/flow/}
\end{enumerate}

\textbf{Outcome:} Two grammars (Timeline, Flow) coexist, both consuming the same kernel. The architecture is validated.

\subsubsection{Phase 2: Extract Packages}

\textbf{Goal:} Split into separate npm packages when publishing/team boundaries justify it.

\textbf{Actions:}
\begin{enumerate}
  \item Create \texttt{packages/kernel/} from \texttt{packages/core/src/kernel/}
  \item Create \texttt{packages/timeline/} from \texttt{packages/core/src/timeline/}
  \item Create \texttt{packages/flow/} from \texttt{packages/core/src/flow/}
  \item Update \texttt{packages/cli/} and \texttt{packages/mcp/} to depend on the new packages
  \item Publish as \texttt{@diagram-compiler/kernel}, \texttt{@diagram-compiler/timeline}, etc.
  \item Deprecate the old \texttt{@timeline-compiler/core} package (or keep as umbrella re-export)
\end{enumerate}

\textbf{Trigger:} Phase 2 is triggered when:
\begin{itemize}
  \item A third grammar is added (justifies the extraction overhead)
  \item OR external contributors need to develop grammars independently
  \item OR the monorepo becomes unwieldy
\end{itemize}

\textbf{Outcome:} Clean package boundaries, independent versioning, clear ownership.

\subsubsection{Phase 3: Grammar Marketplace (Future)}

\textbf{Goal:} Enable community-contributed grammars outside the core monorepo.

\textbf{Actions:}
\begin{enumerate}
  \item Publish kernel with stable API guarantees (semantic versioning)
  \item Document grammar authoring guide
  \item Create template repository for new grammars
  \item Establish discovery mechanism (registry, awesome-list)
\end{enumerate}

\textbf{Outcome:} Third-party grammars that interoperate with the kernel and can be used in compositions.

\subsection{Package Naming}

\begin{center}
\small
\begin{tabular}{ll}
\toprule
\textbf{Package} & \textbf{npm Name} \\
\midrule
Kernel & \texttt{@diagram-compiler/kernel} \\
Timeline Grammar & \texttt{@diagram-compiler/timeline} \\
Flow Grammar & \texttt{@diagram-compiler/flow} \\
Graph Grammar & \texttt{@diagram-compiler/graph} \\
Comparison Grammar & \texttt{@diagram-compiler/comparison} \\
Stat Grammar & \texttt{@diagram-compiler/stat} \\
Composition & \texttt{@diagram-compiler/composition} \\
CLI & \texttt{@diagram-compiler/cli} \\
MCP Server & \texttt{@diagram-compiler/mcp} \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Note:} The existing \texttt{@timeline-compiler/*} packages may be retained as aliases or deprecated with migration guidance.

\subsection{Versioning Strategy}

\begin{description}
  \item[Kernel] Follows semantic versioning strictly. Breaking changes require major version bump.
  
  \item[Grammars] Version independently. May have different version numbers.
  
  \item[Composition] Version tracks the highest grammar version it supports.
  
  \item[CLI/MCP] Version tracks the kernel version.
\end{description}

\paragraph{Compatibility matrix.}
A grammar at version X is compatible with kernel versions [X, X+1). Major kernel versions may break grammars; grammars must be updated.

\subsection{Monorepo Tooling}

The monorepo uses:

\begin{description}
  \item[pnpm workspaces] For package management and linking
  \item[TypeScript project references] For incremental builds and cross-package type checking
  \item[Changesets] For versioning and changelog generation
  \item[Turborepo or Nx] (optional) For build caching and task orchestration
\end{description}

\subsection{CI/CD Considerations}

\begin{itemize}
  \item \textbf{Test matrix:} All grammars tested against kernel changes
  \item \textbf{Publish order:} Kernel publishes before grammars
  \item \textbf{Golden images:} Per-grammar golden test suites
  \item \textbf{Cross-platform:} macOS, Linux, Windows in CI
\end{itemize}
